Space as a Scaffold for Rotational Generalisation of Abstract Concepts

Version 1

Abstract

Abstract

Learning invariances allows us to generalise. In the visual modality, invariant representations allow us to recognise objects despite translations or rotations in physical space. However, how we learn the invariances that allow us to generalise abstract patterns of sensory data (“concepts”) is a longstanding puzzle. Here, we study how humans generalise relational patterns in stimulation sequences that are defined by either transitions on a nonspatial two-dimensional feature manifold, or by transitions in physical space. We measure rotational generalisation, that is the ability to recognise concepts even when their corresponding transition vectors are rotated. We find that humans naturally generalise to rotated exemplars when stimuli are defined in physical space, but not when they are defined as positions on a nonspatial feature manifold. However, if participants are first pre-trained to map auditory or visual features to spatial locations, then rotational generalisation becomes possible even in nonspatial domains. These results imply that space acts as a scaffold for learning more abstract conceptual invariances.

Introduction

To recognise objects and events in the natural world, humans form mental representations that are invariant to transformation. The existence of invariant representations allows entities to be recognised and categorised despite changes in their surface properties, which is called “generalisation”. The formation of invariances has been most extensively studied in the case of visual object recognition. For example, we have no trouble recognising a teapot that is moved to a new location (translated), tipped on its side (rotated) or viewed from afar (rescaled). How we do so has provoked diverse theories based on assembly from geometric primitives [1,2], associative learning [3,4], and function approximation in deep networks [5].

A core problem in cognitive science, however, is how we form invariances over entities that are defined by more abstract relational properties. Here, we use the term “concepts” to refer to objects or events that are defined by shared relations among features that may unfold in space, or time, or both, in any modality [6]. For example, the concept of a “tree” implies an entity whose structure is defined by a nested hierarchy, whether this is a physical object whose parts are arranged in space (such as an oak tree in a forest) or a more abstract data structure (such as a family tree or taxonomic tree). The concept of a “ring” implies an entity whose features are arranged cyclically, whether a physical ring (worn on the finger), the (circular) temporal pattern of tones in a peal of bells, or the periodicity in the passage of the seasons [7]. The human ability to readily form invariances over abstract concepts remains a puzzle for both cognitive and neural scientists hoping to understand neural computations, and a challenge for AI researchers wishing to build intelligent agents.

One prominent theory argues that we learn invariant concepts because of the way the brain represents physical space [8–12]. This argument states that neurons coding for positions in either egocentric (viewer-centred) or allocentric (world-centred) space can be recycled to represent locations in more abstract spaces, defined by continuous variation in features (e.g., red to blue, quiet to loud). This theory is backed up by proof-of-concept computational simulations [13], and by findings that brain regions thought to be critical for spatial cognition in mammals (such as the hippocampal-entorhinal complex and parietal cortex) exhibit neural codes that are invariant to relational transformations of nonspatial stimuli [14–18]. However, whilst promising, this theory lacks direct empirical evidence. Here, we set out to provide a strong test of the idea that learning about physical space scaffolds conceptual generalisation. Our focus is on the ability to generalise knowledge about the relations among items in a sequence as they are translated or rotated through both spatial and nonspatial domains.

In the four studies described here, participants made category judgments about a sequence of four successive stimuli in either auditory, visual or spatial modalities. In auditory and visual modalities, the stimulus was drawn from a two-dimensional feature manifold (e.g., a bivariate “space” defined by colour and shape in the visual modality or pitch and timbre in the auditory modality). In the spatial modality, each stimulus was a position in physical space (e.g., an x and y coordinates). Concepts were defined by a common pattern of transitions through either feature space or physical space. Our research question concerned the conditions under which concepts could be recognised, even if their corresponding transition vectors had been translated or rotated. We studied generalisation of transition vectors both within the same feature space and to new feature spaces in the same modality.

Our studies measure the tendency to generalise both by translation and rotation. Conceptual translation occurs when feature values are shifted in either dimension, but with no change in their relational pattern. There is already good evidence that nonspatial concepts are represented in a translation invariant format. For example, in the auditory domain, we can recognise “auditory objects” that are translated in feature space (e.g., pitch and timbre). This occurs when we understand the same sentence from different speakers, or identify the same melody played with different musical instruments [19,20]. However, much less is known about the learning of rotational invariances for abstract concepts. In physical space, we readily learn rotation-invariant object representations (allowing us to recognise an upside down teapot), and the computational mechanisms by which we do so have been a major fulcrum of debate in the vision sciences [3,4]. But whether participants can learn rotationally invariant concepts in nonspatial domains, i.e. those that are defined by sequences of visual and auditory features (rather than by locations in physical space, defined in Cartesian or polar coordinates) is not known. In the current study, we first test this, and find that naively, they cannot. Next, turning to our main hypothesis, we then ask if first teaching participants to map nonspatial features to spatial locations (providing a spatial scaffold) allows the learning of rotational invariances, even in nonspatial modalities. We find that it does. This shows that a form of generalisation that is not usually possible for humans becomes possible when their understanding of the concept is “scaffolded” by first learning a corresponding spatial representation. This thus supports the theory that abstract concept learning is linked to our understanding of physical space.

Results

On each trial, participants were presented with a sequence of four auditory, visual or spatial stimuli (a quadruplet) drawn from one of 16 points on a continuously varying 2D (4 x 4) feature manifold. In the visual (auditory) modality, this manifold was respectively defined by two orthogonal and continuously varying visual (auditory) features. In the spatial domain, the 2D feature manifold was defined by positions in physical space, in either Cartesian or polar coordinates (see Fig. 1D, Fig. S1-S2). Each quadruplet was constructed by first sampling a random point on the 2D feature manifold, and then iteratively choosing three further adjacent feature locations to make a sequence of four stimuli. In each experiment, there were three categories (Fig. 1C). Each category was initially defined by a canonical set of transition vectors, which specified the three successive steps on the feature manifold (defining the positions from which stimuli in the quadruplet were sampled). Thus, for example, one set of transition vectors might be defined by compass directions {NE, W, SE}. This would mean that after an initial stimulus was sampled, the second stimulus in the sequence would be the one NE in feature space, and the third W of that, and the fourth SE of that). We define rotational generalisation as the ability to recognise regularities in the sequence transition vectors that are independent of both translation and rotation. Thus, just as an upside-down teapot can still be recognised by the relative spatial relations among its handle, body, and spout, in Exp. 1 we asked whether concepts can be recognised when their associated transition vectors are rotated (e.g., vector sequence {NE, W, SE} on the feature manifold becomes {NW, S, NE} after 90° rotation). Note that in our study, quadruplets are also randomly translated on the manifold by virtue of the variable initial feature selection between trials. We thus make the basic assumption that rotational generalisation also involves translation invariance.

Paradigm.

A. Objects can be perceived as similar despite changes in shape, orientation and features. B. Similarly, can sequences of features be perceived as similar despite great changes in the features that composed them? C. In our experiments, participants had to learn the category associated with quadruplets composed of four stimuli drawn from a 2D feature manifold. To do so, they could use one of two major strategies: tracking changes in a single dimension (1D) or two dimensions (2D). D. Illustration of the feature manifold and transition vectors in the visual modality. Each transition vector defined transitions between cardinally or diagonally adjacent features on a 2D manifold (here, given by colour and spikiness for training and near transfer in the left and middle panels, and by transparency and squareness in the far transfer condition). The transition vectors for each category are shown in their canonical (0°) rotation as blue, orange and green arrows superimposed on each feature manifold. The vectors are shown rotated by 90° in the near and far transfer conditions. Next to each feature manifold is a matrix showing the expected mapping (filled squares) from feature vectors to categories for a participant using the 1D strategies (top and middle) and 2D strategy (bottom). Note that the 2D strategy always leads to effective generalisation (rotated exemplars are mapped on their corresponding categories) even in transfer, as indicated by filled squares on the matrix diagonal, whereas a 1D strategy leads to a different pattern. We use the symbols ∅, ↶ and ★ to denote canonical (0°), 90° rotation and far transfer conditions respectively.

Our basic procedure was as follows. During training (120 trials), participants first learned to assign canonical (0° rotation) quadruplets to one of three categories using a button press, receiving fully informative feedback after each response (see Fig. S3). Then, during test, participants performed a further 210 trials, half of which were identical to training (with feedback) while the other half were transfer trials involving categorisation of quadruplets whose feature transition vectors were rotated by 90°, 180° or 270°. These novel quadruplets were either sampled from the same 2D feature manifold (e.g., colour and spikiness in the visual case; near transfer condition) or a new 2D feature manifold from the same modality (e.g., transparency and squareness; far transfer condition; see Fig. 1D, Fig. S2). Transfer trials received no feedback, allowing us to infer what knowledge was being generalised between training and transfer. Experiments 1-3 were pre-registered at https://osf.io/z9572/registrations.

Concepts defined by spatial locations, but not auditory or visual features, are rotation-invariant

In Exp. 1, we recruited three cohorts of online participants (N = 50 each, see Fig. 2) to perform the task in the auditory, visual and spatial modalities. These conditions differed only in how the feature manifold was defined: e.g., fundamental frequency and modulation frequency for auditory features (Fig. 2A); e.g., spikiness and colour for visual features (Fig. 2E); e.g., horizontal and vertical position for spatial locations (Fig. 2I). Accuracy on training trials for each modality is shown in Fig. 2B, F, J. Participants learned the task well in all three conditions (but better in the spatial modality: intercept β = 2.90 ± 0.19, slope associated with the auditory modality β = -1.91 ± 0.27, p < 0.001, slope associated with the visual modality, β = -1.57 ± 0.26, p < 0.001, mixed logistic regression on the probability of a correct response with participants as random effect). However, our main question was how participants would generalise learning to novel, rotated exemplars of the same concept.

Experiment 1: Quadruplets composed of spatial locations, but not auditory or visual features, are associated with a 2D strategy.

A. In the auditory modality (N = 50), the 2D feature manifold was defined by auditory features. Here, we illustrate with a schematic in which the feature manifold is defined by fundamental frequency and frequency modulation. B. Accuracy on training trials involving the canonical transition vectors (0° rotation, denoted by the symbol ∅). The bold curve and dots represent the group average; lighter curves are individual participants. During training, random models (R) are at chance (33% accuracy; lower dashed line), while idealised 1D and 2D models are at ceiling (100% accuracy; upper dashed lines). C. Model fits to the near transfer responses. The bar plot shows model frequencies in the population and black dots are Bayesian point estimates of the model frequency. The matrices show cross-validated model predictions in the same format at Fig. 1D, except with the degree of fading (light to dark colour) signalling average participant responses in each case. The matrices read as follows: the top row depicts the average behaviour of participants who were best fit by the 1Diu model (using a held out set of responses). The 1Diu model predictions are shown in black for three rotations of the quadruplets (90°, 180°, 270° denoted by the symbols ↶, ↺ and ↷ respectively). The average response matrices of the participants using held-out responses are shown in red. The scatter plot below shows individual likelihoods for the 1D and 2D models (normalised by the R model). Each dot is an individual participant, coded by whether they are best fit by either 1D or 2D models (colour) or the R model (grey). Dashed lines distinguish zones of relative likelihood where participants are best fit by R (bottom left square), 1D (rightmost quadrilateral) or 2D (upper quadrilateral) models. D. Model fits to the far transfer responses, using the same conventions as C. Note that, at transfer, more 1D models are possible because of the differing ways that participants could map between the i and j axes and the ★i and ★j axes during training and far transfer. E. In the visual modality (N = 52), the 2D feature manifold was defined by visual attributes; here we use colour and spikiness as an illustration. F-G-H depict data from the visual modality using conventions from B-C-D. I. In the spatial modality (N = 51), the 2D feature manifold was defined by the spatial location of a star (here a red dot is used for illustration). J-K-L depict data from the spatial modality using conventions from B-C-D.

To test this, we fit a family of quantitative models jointly to the training and transfer trials. To understand the logic of this modelling exercise, it is necessary to consider the alternative strategies that participants may have learned during training. Whereas rotation requires participants to represent both dimensions of the feature manifold (a rotation of 90° is only discernible in 2D), a viable alternative strategy during training is to base categorical decisions on a single feature (e.g., either spikiness or colour but not both). Each quadruplet consists of four adjacent feature locations forming a square on the feature manifold (Fig. 2D) and thus the stimulation sequence comprises two features from each dimension. Thus, for example, if a participant attended only to spikiness, the four stimuli in a quadruplet would be represented as a feature pattern over spikiness levels (such as ABAB or ABBA, where A = more spiky, B = less spiky). During training, participants could learn to map these patterns onto categories, either in a signed fashion (e.g., ABAB maps to one category and ABBA to another), or an unsigned fashion (ABAB and BABA both map to the same category). These strategies would lead to perfect performance during training, but would prevent the learning of rotational invariances. We built models that implemented these one-dimensional strategies, which we call 1Ds and 1Du respectively, and compared them to models that used both dimensions for categorisation (the 2D model) or were simply responding randomly (R models). Each of these models predicts a unique pattern of generalisation (Fig. S4-S5) and only the 2D model predicts that participants will assign rotated objects to the same category as their unrotated counterparts (rotational generalisation). Thus, the principal metrics we report in this study are the fraction of (non-random) subjects classified as 1D vs. 2D on transfer trials, which is a signature of whether the experimental conditions permitted the learning of rotational invariances for quadruplets. We report both fractions of participants (X/X best fit by each model) and Bayes Factors reflecting the relative likelihood of 1D vs. 2D models between conditions.

Consistent with our first pre-registered prediction, Exp. 1 revealed a striking dissociation in rotational generalisation between modalities. For near transfer, all non-random participants in the auditory and visual modality (26/26 and 38/38) learned a 1Du strategy, whereas the vast majority in the spatial modality (35/41) were best fit by a 2D strategy. Bayesian group model comparison confirmed that the frequency of 1D vs 2D models among non-random participants was similar between the auditory and visual modalities (Bayes Factor [BF] = 0.1, “negative” evidence for a difference) but different between the auditory and spatial modalities (BF > 100, “decisive” evidence for a difference) and the visual and spatial modalities (BF > 100; see tables S6-S9 for full results). This implies that the use of a 1D strategy (implying no rotational generalisation) was much more likely than a 2D strategy (implying rotational generalisation) when the manifold was defined by visual or auditory features (e.g., colour and shape or pitch and timbre), but the converse was true when the feature manifold was defined by coordinates in physical space (e.g., horizontal and vertical position).

For far transfer, the results were very similar. In the auditory modality, all non-random participants (26/26) were again best fit by a 1Du strategy, and in the visual modality, most (30/35) were fit by a 1Du strategy, 5/35 by a 1Ds strategy, and none by a 2D strategy (difference between auditory and visual, BF = 0.1). By contrast, in the spatial modality, where far transfer involved remapping from cardinal to polar coordinates or vice versa, almost all non-random participants (29/31) were again best fit by a 2D strategy (both BF > 100 comparing with the auditory and visual modalities, “decisive” evidence for a difference). Behaviour in each modality of Exp. 1 is illustrated in Fig. 2, where we display category assignments under each rotation for participants allocated to distinct model classes on the basis of held out data. Together, these data show definitively that, when categories were characterised by temporal patterns in spatial location (e.g., where transitions in physical space were aligned with those on the feature manifold), participants learned to represent the 2D structure of the concept, and generalised readily to rotated (as well as translated) exemplars. However, when concepts were defined by patterns of nonspatial auditory or visual features, participants learned mappings to each category by relying on a single feature dimension and thus failed to form rotational invariant representations.

Spatial pre-training provides a scaffold for rotational generalisation in the auditory and visual modalities

Exp. 1 shows that rotational generalisation succeeds for spatial concepts but fails for nonspatial concepts. Next, in Exp. 2 we tested our main prediction: that space can be used as a scaffold for rotational generalisation of nonspatial concepts. We recruited three new cohorts of participants (N = 50 each, see Fig. 3) to perform a multi-phase task that unfolded over two successive days. On day 1, participants received 60 pre-training trials in the pre-training modality. These trials matched training trials in Exp. 1 for the corresponding modality (spatial or visual) except that they comprised both canonical (0°) and 90° rotated quadruplets, but not those rotated by 180° or 270° (we included examples of rotated quadruplets in the training set to encourage rotational generalisation, but as shown in Exp. 3, results do not depend on this choice). Subsequently, participants performed 288 trials of a multimodal association task, in which they learned the association between each of the 16 stimuli in the pre-training modality and their corresponding stimulus in a different testing modality, where the corresponding stimulus occupied an equivalent position on the 2D feature manifold (we call this the “mapping task”). The goal of this task was to teach participants correspondences between either spatial and visual, spatial and auditory, or visual and auditory feature manifolds. Then on day 2, after some refresher pre-training and mapping trials, participants performed the same task as in Exp. 1 in the testing modality, again with the exception that training trials also included 90° rotated quadruplets.

Experiment 2: Spatial pre-training triggers the use of a 2D strategy for quadruplets composed of auditory or visual features.

A. Performance on the spatial pre-training task in Exp. 2a (N = 51). For this and all plots below, the bold line is the group average and lighter lines are individual participants. Dashed lines show expected average performance under the corresponding models (labelled). B. Performance in the spatial to auditory mapping task. Chance performance is at 1/16 (dashed line). C. Performance on training trials in the auditory modality for canonical (0°) and 90° rotated quadruplets. During training, random models (R) are at chance (33% accuracy), almost all 1D models are at 50% accuracy, and 1Dis and 2D models are at ceiling (100% accuracy). D. Model fits to the near transfer responses, using the same conventions as Fig. 2. Model predictions are shown in black for two rotations of the quadruplets (180° and 270° denoted by the symbols ↺ and ↷ respectively). E. Model fits to the far transfer responses. F-G-H-I-J read as A-B-C-D-E for Exp. 2b (N = 51), where participants performed a spatial pre-training and a spatial to visual mapping task prior to performing the visual version of the main task. K-L-M-N-O read as A-B-C-D-E for Exp. 2c (N = 50), where participants performed a visual pre-training and a visual to auditory mapping task prior to performing the auditory version of the main task.

Our pre-registered prediction for Exp. 2 was that when (Cartesian) physical space was the pre-training modality, participants would now (in contrast to Exp. 1) learn using a predominantly 2D strategy in both auditory (Exp. 2a) and visual (Exp. 2b) testing modalities. In other words, by learning the association between auditory or visual features and a corresponding spatial location, concepts composed of exclusively nonspatial features could now be generalised over rotations in a way not exhibited by a single participant in Exp. 1a or Exp. 1b. By way of control, however, we predicted that when the pre-training involved (nonspatial) visual features, no such benefit would occur, and participants would fail to show rotation invariance.

This is exactly what we found, for both near and far transfer. In the near transfer condition, with spatial pre-training, 29/37 non-random participants were best fit by a 2D strategy when audition was the testing modality, and 36/40 when vision was the testing modality. By contrast however, participants who underwent visual (rather than spatial) pre-training failed to show a benefit when audition was the testing modality. In fact, most (37/50) were best fit by the random model (see below), with 6/13 non-random participants favouring a 1Ds strategy. We once again calculated Bayes Factors at the group level to assess the reliability of these results. We found that BFs exceeded 100, providing “decisive” evidence that the 2D model was more favoured among the groups with spatial pre-training than that without. Similarly, in the far transfer condition, spatial pre-training allowed 29/38 participants in the auditory modality and 13/16 participants in the visual modality to successfully generalise via a 2D strategy. This was not the case for participants who experienced visual pre-training (again, frequency of 1D vs 2D models between conditions: BF > 100, “decisive”). In other words, spatial pre-training provided an effective scaffold that allowed participants to learn auditory and visual objects in a 2D representational format that permitted generalisation to novel rotated exemplars.

Participants in Exp. 2c performed poorly during training and were more likely to be fit by the random models during transfer than those who performed the same auditory task in Exp. 1a. Indeed, we computed the Bayes Factor quantifying the relative likelihood of the random (R models) vs all other models (1D and 2D models) and found “substantial” evidence in favour of a difference between groups both in near transfer (BF = 7.4) and in far transfer (BF = 2.7). This might seem curious, because Exp. 2 participants had access to more diverse training (on both 0° and 90° quadruplets) as well as the supplementary visual pre-training. Why did Exp. 2c participants struggle with the task? In fact, this phenomenon makes sense, because training with feedback on both 0° and 90° quadruplets effectively invalidates a 1D strategy, because there no longer exists a unique mapping between categories and features in either of the two feature dimensions (note that training performance in Exp. 2c plateaus close to 50%). This lack of a viable 1D strategy during training obliges participants to use a 2D strategy where possible. Because this is only possible with spatial pre-training, in Exp. 2c they revert to random. Whilst this explains what we observed in Exp. 2, it also allows a further prediction: if we remove the 90° rotated quadruplets from pre-training, then participants in the spatial pre-training modality should be somewhat less prone to use a 2D strategy (because 1D is available) whereas participants who undergo visual pre-training should show more 1D behaviour at the expense of the random model. In Exp. 3, we tested and confirmed this prediction.

Exp. 3 involved three new cohorts (N = 50 each, see Fig. 4) and was identical to Exp. 2, except that now pre-training trials consisted exclusively of canonical (0°) quadruplets (although 90° quadruplets were still present when the testing modality was trained on day 2). As predicted, non-random participants who enjoyed spatial pre-training were still prone to use a 2D strategy when audition was the testing modality (16/32 for near transfer and 17/31 for far transfer) as well as when vision was the testing modality (22/30 and 13/19), replicating the findings of Exp. 2. However, compared to Exp. 2, overall more participants relied on 1D strategies. In the auditory modality in Exp. 3a, 16/32 were best fit by a 1D model in the near transfer condition (9/32 1Du and 7/32 1Ds) and 14/31 in the far transfer condition (10/31 1Du and 4/31 1Ds). At the group level, the Bayes Factor confirmed that participants were more likely to be fit by a 1D model in Exp. 3a than Exp. 2a in the auditory modality (frequency of 1D vs 2D models between Exp. 2a and Exp. 3a, near transfer BF = 3.4 “substantial” evidence, far transfer BF = 1.5 “weak” evidence). Similarly, in the visual modality in Exp. 3b, 8/30 were best fit by a 1D model in the near transfer condition (5/30 1Du and 3/30 1Ds) and 6/19 in the far transfer condition (6/19 1Du) (again, frequency of 1D vs 2D models between Exp. 2b and Exp. 3b, near transfer BF = 5.9 “substantial” evidence, far transfer BF = 2.2 “weak” evidence). By contrast, participants who underwent nonspatial (visual) pre-training did not use a 2D strategy (1/30) but rather preferred 1D strategies in both near transfer (13/30 1Du and 16/30 1Ds) and far transfer conditions (10/28 1Du and 17/28 1Ds). Comparing these results with the frequency of 1D vs 2D models in conditions with spatial pre-training (Exp. 3a and 3b), we found that all BFs exceeded 100, providing “decisive” evidence that the 2D model was more favoured among the groups with spatial pre-training than that without.

Experiment 3: Exposure to the canonical events (0°) during spatial pre-training is sufficient to trigger the use of a 2D strategy.

All panels use the same conventions as Fig. 3. A-B-C-D-E. Contrary to Exp. 2a, in Exp. 3a (N = 50), participants were exposed only to canonical events (0°) during the spatial pre-training. F-G-H-I-J. Exp. 3b (N = 51). K-L-M-N-O. Exp. 3c (N = 50).

Thus, these results show that training exclusively on canonical (0°) quadruplets facilitates a 1D strategy, which is expressed more readily than in Exp. 2; but that the 2D strategy is still more likely for participants who underwent spatial pre-training. Thus, spatial pre-training provides a scaffold to learn rotation-invariant representation of auditory and visual concepts even when rotation is never explicitly shown during pre-training. Furthermore, participants are sensitive to the available strategies during pre-training, and use the 1D strategy when possible if they have not learned to associate features with space.

Spatial mapping performance predicts rotational generalisation for nonspatial modalities

Next, we used our data from Exp. 2 and Exp. 3 to study how performance on each phase of our task predicted rotational generalisation in the testing phase (see Fig. 5). For each participant, we created an index of rotation generalisation (2Dness) as the difference in log-likelihood between the best 1D model and the 2D model during near transfer. We found that 2Dness was powerfully predicted by training accuracy (Pearson correlation between 2Dness and training accuracy [r2Dness,TRAINING] = 0.80, p < 0.001) in both Exp. 2 (r2Dness,TRAINING = 0.83, p < 0.001) and Exp. 3 (r2Dness,TRAINING = 0.78, p < 0.001). The fact that training performance is highly correlated with 2Dness implies that participants who solved the training task formed representations that were generalisable in 2D; in other words, very few participants overfit to the training set. Accordingly, participants were poorly captured by an additional model (the R’ model; 4/152 in Exp. 2, 4/151 in Exp. 3), that has perfect performance during training but responds randomly during transfer. Next, we asked whether pre-training and mapping were systematically associated with 2Dness, and assessed their relative importance using partial correlations. Pre-training did explain unique variance in 2Dness after accounting for mapping (correlation between pre-training and 2Dness after partialling out mapping [r2Dness,PRETRAINING - MAPPING] = 0.17, p < 0.01) and vice versa (r2Dness,MAPPING - PRETRAINING = 0.27, p < 0.001). However, 2Dness was better predicted by mapping than by pre-training in Exp. 2a (r2Dness,MAPPING - PRETRAINING = 0.42, p < 0.005 and r2Dness,PRETRAINING - MAPPING = 0.30, p < 0.05), Exp. 2b (r2Dness,MAPPING - PRETRAINING = 0.41, p < 0.005 and r2Dness,PRETRAINING - MAPPING = 0.04, p = 0.81), Exp. 3a (r2Dness,MAPPING - PRETRAINING = 0.32, p < 0.05 and r2Dness,PRETRAINING - MAPPING = 0.06, p = 0.70) and Exp. 3b (r2Dness,MAPPING - PRETRAINING = 0.46, p < 0.001 and r2Dness,PRETRAINING - MAPPING = 0.33, p < 0.05). The strong correlations between the mapping task performances and 2Dness suggest that learning the association between nonspatial and spatial features is the critical step that allows rotational generalisation.

Correlation analysis.

Scatter plots of 2Dness (the difference in likelihood between the best 1D model and the 2D model in the training modality in near transfer) and A. pre-training accuracy, B. mapping accuracy and C. training accuracy in the testing modality in Exp. 2. Dots are individual participants in Exp. 2a (red), 2b (blue) and 2c (brown). D-E-F read as A-B-C for Exp. 3.

We tested and confirmed this prediction in Exp. 4 (see Fig. S7) which repeated Exp. 3 except that spatial pre-training was replaced with a duration-matched filler task (in which the category is defined by the number of stationary blue stars in a sequence). Without spatial pre-training, a sizeable proportion of participants still learned a 2D strategy in both the auditory (9/30 in near transfer, 9/28 in far transfer) and visual (12/19 and 7/19) modality, although the majority relied on a 1D strategy (auditory modality: 4/30 1Du and 17/30 1Ds for near transfer, 4/38 1Du and 15/28 1Ds for far transfer; visual modality: 5/19 1Du and 6/19 1Ds for near transfer, 5/14 1Du and 2/14 1Ds for far transfer). In the auditory modality (Exp. 4a), this can be compared with Exp. 2c, where almost all participants were using a random strategy (frequency of R vs 1D/2D models, BF = 34.0, “strong” evidence), and with Exp. 3c where almost no participants were using a 2D strategy (frequency of 1D vs 2D models, BF = 8.6, “substantial” evidence). Thus, for ~20% participants, the mere exposure to the mapping was sufficient to benefit from the spatial scaffolding effect and actually seeing the quadruplets in the spatial modality was not necessary for them.

Discussion

We studied the conditions under which participants learn rotation- and translation-invariant representations of abstract concepts. We found that participants can generalise conceptual knowledge to novel sequences (quadruplets) defined by rotations of stimulus feature transition vectors, but only if the features were themselves are physical spatial locations (e.g., x, y position; Exp. 1) or if nonspatial attributes had previously been mapped to a physical spatial location in a pre-training task (Exp. 2-4). Thus, an explicit representation of physical space is a “scaffold” that permits objects to be learned in a rotation invariant fashion, and thus allows rotational generalisation. This supports the idea that neural representations of space for a critical substrate for learning abstractions in nonspatial domains [8–10,21].

It is well known that humans learn rotational invariances for visual objects, whose features are organised in physical space. For example, an upside down teapot can be recognised by the relative position of handle, lid and spout. This case mimics our spatial modality condition, where each concept was a pattern of locations in physical space. It is thus perhaps unsurprising that rotational generalisation is possible in this condition. However, we found it striking that participants generalised in such different ways when the features in question were drawn from a nonspatial manifold, in either the visual or auditory domain. In these conditions, participants seemed to have no trouble recognising patterns that were consistently translated in feature space. This is consistent with previous studies that have shown that we can understand language in different accents, or name a familiar tune played at an atypical speed or pitch [22]. However, they did so via a representation that focused on just one of the two possible dimensions, and thus did not permit rotational generalisation. There was thus a clear dissociation between human ability to generalise patterns in physical space and a more abstract feature space.

Next, we showed that spatial pre-training allowed rotational generalisation even for sequences composed of nonspatial features. This implies that the neural representation of space may serve as a “scaffold”, allowing people to visualise and manipulate nonspatial concepts. Whilst we are not aware of previous studies that have tested spatial scaffolding in the way described here, our findings are consistent with the more general idea that space is represented in an overlapping fashion with nonspatial information, such as time or number [23]. For example, sequences with regular spatial geometry are learned more readily than those composed of arbitrary patterns [24]. Our findings also cohere with evidence that visuospatial skills are correlated with a variety of academic competences, especially in STEM subjects such as maths and engineering [25], and that spatial training interventions (such as teaching mental rotation) in educational settings can improve nonspatial abilities, such as calculus grades [26].

The idea that spatial representations form a generalised substrate for cognition – including for coding temporal structure – draws on a long tradition in philosophy [27], cognitive science [11] and neuroscience [8–10]. The precise substrate for this effect is unclear, but it seems likely that neural assemblies activated by physical locations in space (for example, in parietal or medial temporal lobe areas) are recycled for representing nonspatial patterns in data. In our study, thus, the mapping task facilitates this recycling by teaching participants a point-to-point mapping between nonspatial feature combinations and locations in physical space. Indeed, our correlation analysis and Exp. 4 suggested that successfully learning mappings between spatial and nonspatial features was the strongest determinant of rotational generalisation. This mapping task was presented in an egocentric frame of reference defined by the x, y coordinates of the screen. Explicit representations of location in egocentric space in the primate are found in dorsal stream structures such as the posterior parietal cortex [28]. Current deep networks – which successfully categorise lone objects in a natural image but often fail on tests of relational reasoning or scene understanding – may be hampered by their failure to represent space explicitly in this way [10].

All the effects observed in our experiments were consistent across near transfer conditions (rotation of patterns within the same feature space), and far transfer conditions (rotation of patterns within a different feature space, where features are drawn from the same modality). This shows the generality of spatial training for conceptual generalisation. This means that an explicit representation of space might be the substrate for strong forms of transfer observed in humans, such as when we understand the shared meaning between “red, amber, green” at a traffic light and “ready, set, go” before a race. We did not test transfer across modalities; we leave this for future studies.

eLife Assessment

These ingenious and thoughtful studies present important findings concerning how people represent and generalise abstract patterns of sensory data. The issue of generalisation is a core topic in neuroscience and psychology, relevant across a wide range of areas, and the findings will be of interest to researchers across areas in perception, learning, and cognitive science. The findings have the potential to provide compelling support for the outlined account, but there appear other possible explanations, too, that may affect the scope of the findings but could be considered in a revision.

Reviewer #1 (Public Review):

Summary:

This manuscript reports a series of experiments examining category learning and subsequent generalization of stimulus representations across spatial and nonspatial domains. In Experiment 1, participants were first trained to make category judgments about sequences of stimuli presented either in nonspatial auditory or visual modalities (with feature values drawn from a two-dimensional feature manifold, e.g., pitch vs timbre), or in a spatial modality (with feature values defined by positions in physical space, e.g., Cartesian x and y coordinates). A subsequent test phase assessed category judgments for 'rotated' exemplars of these stimuli: i.e., versions in which the transition vectors are rotated in the same feature space used during training (near transfer) or in a different feature space belonging to the same domain (far transfer). Findings demonstrate clearly that representations developed for the spatial domain allow for representational generalization, whereas this pattern is not observed for the nonspatial domains that are tested. Subsequent experiments demonstrate that if participants are first pre-trained to map nonspatial auditory/visual features to spatial locations, then rotational generalization is facilitated even for these nonspatial domains. It is argued that these findings are consistent with the idea that spatial representations form a generalized substrate for cognition: that space can act as a scaffold for learning abstract nonspatial concepts.

Strengths:

I enjoyed reading this manuscript, which is extremely well-written and well-presented. The writing is clear and concise throughout, and the figures do a great job of highlighting the key concepts. The issue of generalization is a core topic in neuroscience and psychology, relevant across a wide range of areas, and the findings will be of interest to researchers across areas in perception and cognitive science. It's also excellent to see that the hypotheses, methods, and analyses were pre-registered.

The experiments that have been run are ingenious and thoughtful; I particularly liked the use of stimulus structures that allow for disentangling of one-dimensional and two-dimensional response patterns. The studies are also well-powered for detecting the effects of interest. The model-based statistical analyses are thorough and appropriate throughout (and it's good to see model recovery analysis too). The findings themselves are clear-cut: I have little doubt about the robustness and replicability of these data.

Weaknesses:

I have only one significant concern regarding this manuscript, which relates to the interpretation of the findings. The findings are taken to suggest that "space may serve as a 'scaffold', allowing people to visualize and manipulate nonspatial concepts" (p13). However, I think the data may be amenable to an alternative possibility. I wonder if it's possible that, for the visual and auditory stimuli, participants naturally tended to attend to one feature dimension and ignore the other - i.e., there may have been a (potentially idiosyncratic) difference in salience between the feature dimensions that led to participants learning the feature sequence in a one-dimensional way (akin to the 'overshadowing' effect in associative learning: e.g., see Mackintosh, 1976, "Overshadowing and stimulus intensity", Animal Learning and Behaviour). By contrast, we are very used to thinking about space as a multidimensional domain, in particular with regard to two-dimensional vertical and horizontal displacements. As a result, one would naturally expect to see more evidence of two-dimensional representation (allowing for rotational generalization) for spatial than nonspatial domains.

In this view, the impact of spatial pre-training and (particularly) mapping is simply to highlight to participants that the auditory/visual stimuli comprise two separable (and independent) dimensions. Once they understand this, during subsequent training, they can learn about sequences on both dimensions, which will allow for a 2D representation and hence rotational generalization - as observed in Experiments 2 and 3. This account also anticipates that mapping alone (as in Experiment 4) could be sufficient to promote a 2D strategy for auditory and visual domains.

This "attention to dimensions" account has some similarities to the "spatial scaffolding" idea put forward in the article, in arguing that experience of how auditory/visual feature manifolds can be translated into a spatial representation helps people to see those domains in a way that allows for rotational generalization. Where it differs is that it does not propose that space provides a *scaffold* for the development of the nonspatial representations, i.e., that people represent/learn the nonspatial information in a spatial format, and this is what allows them to manipulate nonspatial concepts. Instead, the "attention to dimensions" account anticipates that ANY manipulation that highlights to participants the separable-dimension nature of auditory/visual stimuli could facilitate 2D representation and hence rotational generalization. For example, explicit instruction on how the stimuli are constructed may be sufficient, or pre-training of some form with each dimension separately, before they are combined to form the 2D stimuli.

I'd be interested to hear the authors' thoughts on this account - whether they see it as an alternative to their own interpretation, and whether it can be ruled out on the basis of their existing data.

Reviewer #2 (Public Review):

Summary:

In this manuscript, L&S investigates the important general question of how humans achieve invariant behavior over stimuli belonging to one category given the widely varying input representation of those stimuli and more specifically, how they do that in arbitrary abstract domains. The authors start with the hypothesis that this is achieved by invariance transformations that observers use for interpreting different entries and furthermore, that these transformations in an arbitrary domain emerge with the help of the transformations (e.g. translation, rotation) within the spatial domain by using those as "scaffolding" during transformation learning. To provide the missing evidence for this hypothesis, L&S used behavioral category learning studies within and across the spatial, auditory, and visual domains, where rotated and translated 4-element token sequences had to be learned to categorize and then the learned transformation had to be applied in new feature dimensions within the given domain. Through single- and multiple-day supervised training and unsupervised tests, L&S demonstrated by standard computational analyses that in such setups, space and spatial transformations can, indeed, help with developing and using appropriate rotational mapping whereas the visual domain cannot fulfill such a scaffolding role.

Strengths:

The overall problem definition and the context of spatial mapping-driven solution to the problem is timely. The general design of testing the scaffolding effect across different domains is more advanced than any previous attempts clarifying the relevance of spatial coding to any other type of representational codes. Once the formulation of the general problem in a specific scientific framework is done, the following steps are clearly and logically defined and executed. The obtained results are well interpretable, and they could serve as a good stepping stone for deeper investigations. The analytical tools used for the interpretations are adequate. The paper is relatively clearly written.

Weaknesses:

Some additional effort to clarify the exact contribution of the paper, the link between analyses and the claims of the paper, and its link to previous proposals would be necessary to better assess the significance of the results and the true nature of the proposed mechanism of abstract generalization.

1. Insufficient conceptual setup: The original theoretical proposal (the Tolman-Eichenbaum-Machine, Whittington et al., Cell 2020) that L&S relate their work to proposes that just as in the case of memory for spatial navigation, humans and animals create their flexible relational memory system of any abstract representation by a conjunction code that combines on the one hand, sensory representation and on the other hand, a general structural representation or relational transformation. The TEM also suggests that the structural representation could contain any graph-interpretable spatial relations, albeit in their demonstration 2D neighbor relations were used. The goal of L&S's paper is to provide behavioral evidence for this suggestion by showing that humans use representational codes that are invariant to relational transformations of non-spatial abstract stimuli and moreover, that humans obtain these invariances by developing invariance transformers with the help of available spatial transformers. To obtain such evidence, L&S use the rotational transformation. However, the actual procedure they use actually solved an alternative task: instead of interrogating how humans develop generalizations in abstract spaces, they demonstrated that if one defines rotation in an abstract feature space embedded in a visual or auditory modality that is similar to the 2D space (i.e. has two independent dimensions that are clearly segregable and continuous), humans cannot learn to apply rotation of 4-piece temporal sequences in those spaces while they can do it in 2D space, and with co-associating a one-to-one mapping between locations in those feature spaces with locations in the 2D space an appropriate shaping mapping training will lead to the successful application of rotation in the given task (and in some other feature spaces in the given domain). While this is an interesting and challenging demonstration, it does not shed light on how humans learn and generalize, only that humans CAN do learning and generalization in this, highly constrained scenario. This result is a demonstration of how a stepwise learning regiment can make use of one structure for mapping a complex input into a desired output. The results neither clarify how generalizations would develop in abstract spaces nor the question of whether this generalization uses transformations developed in the abstract space. The specific training procedure ensures success in the presented experiments but the availability and feasibility of an equivalent procedure in a natural setting is a crucial part of validating the original claim and that has not been done in the paper.

2. Missing controls: The asymptotic performance in experiment 1 after training in the three tasks was quite different in the three tasks (intercepts 2.9, 1.9, 1.6 for spatial, visual, and auditory, respectively; p. 5. para. 1, Fig 2BFJ). It seems that the statement "However, our main question was how participants would generalise learning to novel, rotated exemplars of the same concept." assumes that learning and generalization are independent. Wouldn't it be possible, though, that the level of generalization depends on the level of acquiring a good representation of the "concept" and after obtaining an adequate level of this knowledge, generalization would kick in without scaffolding? If so, a missing control is to equate the levels of asymptotic learning and see whether there is a significant difference in generalization. A related issue is that we have no information on what kind of learning in the three different domains was performed, albeit we probably suspect that in space the 2D representation was dominant while in the auditory and visual domains not so much. Thus, a second missing piece of evidence is the model-fitting results of the ⦰ condition that would show which way the original sequences were encoded (similar to Fig 2 CGK and DHL). If the reason for lower performance is not individual stimulus difficulty but the natural tendency to encode the given stimulus type by a combo of random + 1D strategy that would clarify that the result of the cross-training is, indeed, transferring the 2D-mapping strategy.

Reviewer #3 (Public Review):

Summary:

Pesnot Lerousseau and Summerfield aimed to explore how humans generalize abstract patterns of sensory data (concepts), focusing on whether and how spatial representations may facilitate the generalization of abstract concepts (rotational invariance). Specifically, the authors investigated whether people can recognize rotated sequences of stimuli in both spatial and nonspatial domains and whether spatial pre-training and multi-modal mapping aid in this process.

Strengths:

The study innovatively examines a relatively underexplored but interesting area of cognitive science, the potential role of spatial scaffolding in generalizing sequences. The experimental design is clever and covers different modalities (auditory, visual, spatial), utilizing a two-dimensional feature manifold. The findings are backed by strong empirical data, good data analysis, and excellent transparency (including preregistration) adding weight to the proposition that spatial cognition can aid abstract concept generalization.

Weaknesses:

The examples used to motivate the study (such as "tree" = oak tree, family tree, taxonomic tree) may not effectively represent the phenomena being studied, possibly confusing linguistic labels with abstract concepts. This potential confusion may also extend to doubts about the real-life applicability of the generalizations observed in the study and raises questions about the nature of the underlying mechanism being proposed.

Next, the study does not explore whether scaffolding effects could be observed with other well-learned domains, leaving open the question of whether spatial representations are uniquely effective or simply one instance of a familiar 2D space, again questioning the underlying mechanism.

Further doubt on the underlying mechanism is cast by the possibility that the observed correlation between mapping task performance and the adoption of a 2D strategy may reflect general cognitive engagement rather than the spatial nature of the task. Similarly, the surprising finding that a significant number of participants benefited from spatial scaffolding without seeing spatial modalities may further raise questions about the interpretation of the scaffolding effect, pointing towards potential alternative interpretations, such as shifts in attention during learning induced by pre-training without changing underlying abstract conceptual representations.

Conclusions:

The authors successfully demonstrate that spatial training can enhance the ability to generalize in nonspatial domains, particularly in recognizing rotated sequences. The results for the most part support their conclusions, showing that spatial representations can act as a scaffold for learning more abstract conceptual invariances. However, the study leaves room for further investigation into whether the observed effects are unique to spatial cognition or could be replicated with other forms of well-established knowledge, as well as further clarifications of the underlying mechanisms.

Impact:

The study's findings are likely to have a valuable impact on cognitive science, particularly in understanding how abstract concepts are learned and generalized. The methods and data can be useful for further research, especially in exploring the relationship between spatial cognition and abstract conceptualization. The insights could also be valuable for AI research, particularly in improving models that involve abstract pattern recognition and conceptual generalization.

In summary, the paper contributes valuable insights into the role of spatial cognition in learning abstract concepts, though it invites further research to explore the boundaries and specifics of this scaffolding effect.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

These ingenious and thoughtful studies present important findings concerning how people represent and generalise abstract patterns of sensory data. The issue of generalisation is a core topic in neuroscience and psychology, relevant across a wide range of areas, and the findings will be of interest to researchers across areas in perception, learning, and cognitive science. The findings have the potential to provide compelling support for the outlined account, but there appear other possible explanations, too, that may affect the scope of the findings but could be considered in a revision.

Thank you for sending the feedback from the three peer reviewers regarding our paper. Please find below our detailed responses addressing the reviewers' comments. We have incorporated these suggestions into the paper and provided explanations for the modifications made.

We have specifically addressed the point of uncertainty highlighted in eLife's editorial assessment, which concerned alternative explanations for the reported effect. In response to Reviewer #1, we have clarified how Exp. 2c and Exp. 3c address the potential alternative explanation related to "attention to dimensions." Further, we present a supplementary analysis to account for differences in asymptotic learning, as noted by Reviewer #2. We have also clarified how our control experiments address effects associated with general cognitive engagement in the task. Lastly, we have further clarified the conceptual foundation of our paper, addressing concerns raised by Reviewers #2 and #3.

Reviewer #1 (Public Review):

Summary:

This manuscript reports a series of experiments examining category learning and subsequent generalization of stimulus representations across spatial and nonspatial domains. In Experiment 1, participants were first trained to make category judgments about sequences of stimuli presented either in nonspatial auditory or visual modalities (with feature values drawn from a two-dimensional feature manifold, e.g., pitch vs timbre), or in a spatial modality (with feature values defined by positions in physical space, e.g., Cartesian x and y coordinates). A subsequent test phase assessed category judgments for 'rotated' exemplars of these stimuli: i.e., versions in which the transition vectors are rotated in the same feature space used during training (near transfer) or in a different feature space belonging to the same domain (far transfer). Findings demonstrate clearly that representations developed for the spatial domain allow for representational generalization, whereas this pattern is not observed for the nonspatial domains that are tested. Subsequent experiments demonstrate that if participants are first pre-trained to map nonspatial auditory/visual features to spatial locations, then rotational generalization is facilitated even for these nonspatial domains. It is argued that these findings are consistent with the idea that spatial representations form a generalized substrate for cognition: that space can act as a scaffold for learning abstract nonspatial concepts.

Strengths:

I enjoyed reading this manuscript, which is extremely well-written and well-presented. The writing is clear and concise throughout, and the figures do a great job of highlighting the key concepts. The issue of generalization is a core topic in neuroscience and psychology, relevant across a wide range of areas, and the findings will be of interest to researchers across areas in perception and cognitive science. It's also excellent to see that the hypotheses, methods, and analyses were pre-registered.

The experiments that have been run are ingenious and thoughtful; I particularly liked the use of stimulus structures that allow for disentangling of one-dimensional and two-dimensional response patterns. The studies are also well-powered for detecting the effects of interest. The model-based statistical analyses are thorough and appropriate throughout (and it's good to see model recovery analysis too). The findings themselves are clear-cut: I have little doubt about the robustness and replicability of these data.

Weaknesses:

I have only one significant concern regarding this manuscript, which relates to the interpretation of the findings. The findings are taken to suggest that "space may serve as a 'scaffold', allowing people to visualize and manipulate nonspatial concepts" (p13). However, I think the data may be amenable to an alternative possibility. I wonder if it's possible that, for the visual and auditory stimuli, participants naturally tended to attend to one feature dimension and ignore the other - i.e., there may have been a (potentially idiosyncratic) difference in salience between the feature dimensions that led to participants learning the feature sequence in a one-dimensional way (akin to the 'overshadowing' effect in associative learning: e.g., see Mackintosh, 1976, "Overshadowing and stimulus intensity", Animal Learning and Behaviour). By contrast, we are very used to thinking about space as a multidimensional domain, in particular with regard to two-dimensional vertical and horizontal displacements. As a result, one would naturally expect to see more evidence of two-dimensional representation (allowing for rotational generalization) for spatial than nonspatial domains.

In this view, the impact of spatial pre-training and (particularly) mapping is simply to highlight to participants that the auditory/visual stimuli comprise two separable (and independent) dimensions. Once they understand this, during subsequent training, they can learn about sequences on both dimensions, which will allow for a 2D representation and hence rotational generalization - as observed in Experiments 2 and 3. This account also anticipates that mapping alone (as in Experiment 4) could be sufficient to promote a 2D strategy for auditory and visual domains.

This "attention to dimensions" account has some similarities to the "spatial scaffolding" idea put forward in the article, in arguing that experience of how auditory/visual feature manifolds can be translated into a spatial representation helps people to see those domains in a way that allows for rotational generalization. Where it differs is that it does not propose that space provides a scaffold for the development of the nonspatial representations, i.e., that people represent/learn the nonspatial information in a spatial format, and this is what allows them to manipulate nonspatial concepts. Instead, the "attention to dimensions" account anticipates that ANY manipulation that highlights to participants the separable-dimension nature of auditory/visual stimuli could facilitate 2D representation and hence rotational generalization. For example, explicit instruction on how the stimuli are constructed may be sufficient, or pre-training of some form with each dimension separately, before they are combined to form the 2D stimuli.

I'd be interested to hear the authors' thoughts on this account - whether they see it as an alternative to their own interpretation, and whether it can be ruled out on the basis of their existing data.

We thank the Reviewer for their comments. We agree with the Reviewer that the “attention to dimensions” hypothesis is an interesting alternative explanation. However, we believe that the results of our control experiments Exp. 2c and Exp. 3c are incompatible with this alternative explanation.

In Exp. 2c, participants are pre-trained in the visual modality and then tested in the auditory modality. In the multimodal association task, participants have to associate the auditory stimuli and the visual stimuli: on each trial, they hear a sound and then have to click on the corresponding visual stimulus. It is thus necessary to pay attention to both auditory dimensions and both visual dimensions to perform the task. To give an example, the task might involve mapping the fundamental frequency and the amplitude modulation of the auditory stimulus to the colour and the shape of the visual stimulus, respectively. If participants pay attention to only one dimension, this would lead to a maximum of 25% accuracy on average (because they would be at chance on the other dimension, with four possible options). We observed that 30/50 participants reached an accuracy > 50% in the multimodal association task in Exp. 2c. This means that we know for sure that at least 60% of the participants paid attention to both dimensions of the stimuli. Nevertheless, there was a clear difference between participants that received a visual pre-training (Exp. 2c) and those who received a spatial pre-training (Exp. 2a) (frequency of 1D vs 2D models between conditions, BF > 100 in near transfer and far transfer). In fact, only 3/50 participants were best fit by a 2D model when vision was the pre-training modality compared to 29/50 when space was the pre-training modality. Thus, the benefit of the spatial pre-training cannot be due solely to a shift in attention toward both dimensions.

This effect was replicated in Exp. 3c. Similarly, 33/48 participants reached an accuracy > 50% in the multimodal association task in Exp. 3c, meaning that we know for sure that at least 68% of the participants actually paid attention to both dimensions of the stimuli. Again, there was a clear difference between participants who received a visual pre-training (frequency of 1D vs 2D models between conditions, Exp. 3c) and those who received a spatial pre-training (Exp. 3a) (BF > 100 in near transfer and far transfer).

Thus, we believe that the alternative explanation raised by the Reviewer is not supported by our data. We have added a paragraph in the discussion:

“One alternative explanation of this effect could be that the spatial pre-training encourages participants to attend to both dimensions of the non-spatial stimuli. By contrast, pretraining in the visual or auditory domains (where multiple dimensions of a stimulus may be relevant less often naturally) encourages them to attend to a single dimension. However, data from our control experiments Exp. 2c and Exp. 3c, are incompatible with this explanation. Around ~65% of the participants show a level of performance in the multimodal association task (>50%) which could only be achieved if they were attending to both dimensions (performance attending to a single dimension would yield 25% and chance performance is at 6.25%). This suggests that participants are attending to both dimensions even in the visual and auditory mapping case.”

Reviewer #2 (Public Review):

Summary:

In this manuscript, L&S investigates the important general question of how humans achieve invariant behavior over stimuli belonging to one category given the widely varying input representation of those stimuli and more specifically, how they do that in arbitrary abstract domains. The authors start with the hypothesis that this is achieved by invariance transformations that observers use for interpreting different entries and furthermore, that these transformations in an arbitrary domain emerge with the help of the transformations (e.g. translation, rotation) within the spatial domain by using those as "scaffolding" during transformation learning. To provide the missing evidence for this hypothesis, L&S used behavioral category learning studies within and across the spatial, auditory, and visual domains, where rotated and translated 4-element token sequences had to be learned to categorize and then the learned transformation had to be applied in new feature dimensions within the given domain. Through single- and multiple-day supervised training and unsupervised tests, L&S demonstrated by standard computational analyses that in such setups, space and spatial transformations can, indeed, help with developing and using appropriate rotational mapping whereas the visual domain cannot fulfill such a scaffolding role.

Strengths:

The overall problem definition and the context of spatial mapping-driven solution to the problem is timely. The general design of testing the scaffolding effect across different domains is more advanced than any previous attempts clarifying the relevance of spatial coding to any other type of representational codes. Once the formulation of the general problem in a specific scientific framework is done, the following steps are clearly and logically defined and executed. The obtained results are well interpretable, and they could serve as a good stepping stone for deeper investigations. The analytical tools used for the interpretations are adequate. The paper is relatively clearly written.

Weaknesses:

Some additional effort to clarify the exact contribution of the paper, the link between analyses and the claims of the paper, and its link to previous proposals would be necessary to better assess the significance of the results and the true nature of the proposed mechanism of abstract generalization.

(1) Insufficient conceptual setup: The original theoretical proposal (the Tolman-Eichenbaum-Machine, Whittington et al., Cell 2020) that L&S relate their work to proposes that just as in the case of memory for spatial navigation, humans and animals create their flexible relational memory system of any abstract representation by a conjunction code that combines on the one hand, sensory representation and on the other hand, a general structural representation or relational transformation. The TEM also suggests that the structural representation could contain any graph-interpretable spatial relations, albeit in their demonstration 2D neighbor relations were used. The goal of L&S's paper is to provide behavioral evidence for this suggestion by showing that humans use representational codes that are invariant to relational transformations of non-spatial abstract stimuli and moreover, that humans obtain these invariances by developing invariance transformers with the help of available spatial transformers. To obtain such evidence, L&S use the rotational transformation. However, the actual procedure they use actually solved an alternative task: instead of interrogating how humans develop generalizations in abstract spaces, they demonstrated that if one defines rotation in an abstract feature space embedded in a visual or auditory modality that is similar to the 2D space (i.e. has two independent dimensions that are clearly segregable and continuous), humans cannot learn to apply rotation of 4-piece temporal sequences in those spaces while they can do it in 2D space, and with co-associating a one-to-one mapping between locations in those feature spaces with locations in the 2D space an appropriate shaping mapping training will lead to the successful application of rotation in the given task (and in some other feature spaces in the given domain). While this is an interesting and challenging demonstration, it does not shed light on how humans learn and generalize, only that humans CAN do learning and generalization in this, highly constrained scenario. This result is a demonstration of how a stepwise learning regiment can make use of one structure for mapping a complex input into a desired output. The results neither clarify how generalizations would develop in abstract spaces nor the question of whether this generalization uses transformations developed in the abstract space. The specific training procedure ensures success in the presented experiments but the availability and feasibility of an equivalent procedure in a natural setting is a crucial part of validating the original claim and that has not been done in the paper.

We thank the Reviewer for their detailed comments on our manuscript. We reply to the three main points in turn.

First, concerning the conceptual grounding of our work, we would point out that the TEM model (Whittington et al., 2020), however interesting, is not our theoretical starting point. Rather, as we hope the text and references make clear, we ground our work in theoretical work from the 1990/2000s proposing that space acts as a scaffold for navigating abstract spaces (such as Gärdenfors, 2000). We acknowledge that the TEM model and other experimental work on the implication of the hippocampus, the entorhinal cortex and the parietal cortex in relational transformations of nonspatial stimuli provide evidence for this general theory. However, our work is designed to test a more basic question: whether there is behavioural evidence that space scaffolds learning in the first place. To achieve this, we perform behavioural experiments with causal manipulation (spatial pre-training vs no spatial pre-training) have the potential to provide such direct evidence. This is why we claim that:

“This theory is backed up by proof-of-concept computational simulations [13], and by findings that brain regions thought to be critical for spatial cognition in mammals (such as the hippocampal-entorhinal complex and parietal cortex) exhibit neural codes that are invariant to relational transformations of nonspatial stimuli. However, whilst promising, this theory lacks direct empirical evidence. Here, we set out to provide a strong test of the idea that learning about physical space scaffolds conceptual generalisation.“

Second, we agree with the Reviewer that we do not provide an explicit model for how generalisation occurs, and how precisely space acts as a scaffold for building representations and/or applying the relevant transformations to non-spatial stimuli to solve our task. Rather, we investigate in our Exp. 2-4 which aspects of the training are necessary for rotational generalisation to happen (and conclude that a simple training with the multimodal association task is sufficient for ~20% participants). We now acknowledge in the discussion the fact that we do not provide an explicit model and leave that for future work:

“We acknowledge that our study does not provide a mechanistic model of spatial scaffolding but rather delineate which aspects of the training are necessary for generalisation to happen.”

Finally, we also agree with the Reviewer that our task is non-naturalistic. As is common in experimental research, one must sacrifice the naturalistic elements of the task in exchange for the control and the absence of prior knowledge of the participants. We have decided to mitigate as possible the prior knowledge of the participants to make sure that our task involved learning a completely new task and that the pre-training was really causing the better learning/generalisation. The effects we report are consistent across the experiments so we feel confident about them but we agree with the Reviewer that an external validation with more naturalistic stimuli/tasks would be a nice addition to this work. We have included a sentence in the discussion:

“All the effects observed in our experiments were consistent across near transfer conditions (rotation of patterns within the same feature space), and far transfer conditions (rotation of patterns within a different feature space, where features are drawn from the same modality). This shows the generality of spatial training for conceptual generalisation. We did not test transfer across modalities nor transfer in a more natural setting; we leave this for future studies.”

(2) Missing controls: The asymptotic performance in experiment 1 after training in the three tasks was quite different in the three tasks (intercepts 2.9, 1.9, 1.6 for spatial, visual, and auditory, respectively; p. 5. para. 1, Fig 2BFJ). It seems that the statement "However, our main question was how participants would generalise learning to novel, rotated exemplars of the same concept." assumes that learning and generalization are independent. Wouldn't it be possible, though, that the level of generalization depends on the level of acquiring a good representation of the "concept" and after obtaining an adequate level of this knowledge, generalization would kick in without scaffolding? If so, a missing control is to equate the levels of asymptotic learning and see whether there is a significant difference in generalization. A related issue is that we have no information on what kind of learning in the three different domains was performed, albeit we probably suspect that in space the 2D representation was dominant while in the auditory and visual domains not so much. Thus, a second missing piece of evidence is the model-fitting results of the ⦰ condition that would show which way the original sequences were encoded (similar to Fig 2 CGK and DHL). If the reason for lower performance is not individual stimulus difficulty but the natural tendency to encode the given stimulus type by a combo of random + 1D strategy that would clarify that the result of the cross-training is, indeed, transferring the 2D-mapping strategy.

We agree with the Reviewer that a good further control is to equate performance during training. Thus, we have run a complementary analysis where we select only the participants that reach > 90% accuracy in the last block of training in order to equate asymptotic performance after training in Exp. 1. The results (see Author response image 1) replicates the results that we report in the main text: there is a large difference between groups (relative likelihood of 1D vs. 2D models, all BF > 100 in favour of a difference between the auditory and the spatial modalities, between the visual and the spatial modalities, in both near and far transfer, “decisive” evidence). We prefer not to include this figure in the paper for clarity, and because we believe this result is expected given the fact that 0/50 and 0/50 of the participants in the auditory and visual condition used a 2D strategy – thus, selecting subgroups of these participants cannot change our conclusions.

Results of Exp.

1 when selecting participants that reached > 90% accuracy in the last block of training. Captions are the same as Figure 2 of the main text.

Second, the Reviewer suggested that we run the model fitting analysis only on the ⦰ condition (training) in Exp. 1 to reveal whether participants use a 1D or a 2D strategy already during training. Unfortunately, we cannot provide the model fits only in the ⦰ condition in Exp. 1 because all models make the same predictions for this condition (see Fig S4). However, note that this is done by design: participants were free to apply whatever strategy they want during training; we then used the generalisation phase with the rotated stimuli precisely to reveal this strategy. Further, we do believe that the strategy used by the participants during training and the strategy during transfer are the same, partly because – starting from block #4 – participants have no idea whether the current trial is a training trial or a transfer trial, as both trial types are randomly interleaved with no cue signalling the trial type. We have made this clear in the methods:

“They subsequently performed 105 trials (with trialwise feedback) and 105 transfer trials including rotated and far transfer quadruplets (without trialwise feedback) which were presented in mixed blocks of 30 trials. Training and transfer trials were randomly interleaved, and no clue indicated whether participants were currently on a training trial or a transfer trial before feedback (or absence of feedback in case of a transfer trial).”

Reviewer #3 (Public Review):

Summary:

Pesnot Lerousseau and Summerfield aimed to explore how humans generalize abstract patterns of sensory data (concepts), focusing on whether and how spatial representations may facilitate the generalization of abstract concepts (rotational invariance). Specifically, the authors investigated whether people can recognize rotated sequences of stimuli in both spatial and nonspatial domains and whether spatial pre-training and multi-modal mapping aid in this process.

Strengths:

The study innovatively examines a relatively underexplored but interesting area of cognitive science, the potential role of spatial scaffolding in generalizing sequences. The experimental design is clever and covers different modalities (auditory, visual, spatial), utilizing a two-dimensional feature manifold. The findings are backed by strong empirical data, good data analysis, and excellent transparency (including preregistration) adding weight to the proposition that spatial cognition can aid abstract concept generalization.

Weaknesses:

The examples used to motivate the study (such as "tree" = oak tree, family tree, taxonomic tree) may not effectively represent the phenomena being studied, possibly confusing linguistic labels with abstract concepts. This potential confusion may also extend to doubts about the real-life applicability of the generalizations observed in the study and raises questions about the nature of the underlying mechanism being proposed.

We thank the Reviewer for their comments. We agree that we could have explained ore clearly enough how these examples motivate our study. The similarity between “oak tree” and “family tree” is not just the verbal label. Rather, it is the arrangement of the parts (nodes and branches) in a nested hierarchy. Oak trees and family trees share the same relational structure. The reason that invariance is relevant here is that the similarity in relational structure is retained under rigid body transformations such as rotation or translation. For example, an upside-down tree can still be recognised as a tree, just as a family tree can be plotted with the oldest ancestors at either top or bottom. Similarly, in our study, the quadruplets are defined by the relations between stimuli: all quadruplets use the same basic stimuli, but the categories are defined by the relations between successive stimuli. In our task, generalising means recognising that relations between stimuli are the same despite changes in the surface properties (for example in far transfer). We have clarify that in the introduction:

“For example, the concept of a “tree” implies an entity whose structure is defined by a nested hierarchy, whether this is a physical object whose parts are arranged in space (such as an oak tree in a forest) or a more abstract data structure (such as a family tree or taxonomic tree). [...] Despite great changes in the surface properties of oak trees, family trees and taxonomic trees, humans perceive them as different instances of a more abstract concept defined by the same relational structure.”

Next, the study does not explore whether scaffolding effects could be observed with other well-learned domains, leaving open the question of whether spatial representations are uniquely effective or simply one instance of a familiar 2D space, again questioning the underlying mechanism.

We would like to mention that Reviewer #2 had a similar comment. We agree with both Reviewers that our task is non-naturalistic. As is common in experimental research, one must sacrifice the naturalistic elements of the task in exchange for the control and the absence of prior knowledge of the participants. We have decided to mitigate as possible the prior knowledge of the participants to make sure that our task involved learning a completely new task and that the pre-training was really causing the better learning/generalisation. The effects we report are consistent across the experiments so we feel confident about them but we agree with the Reviewer that an external validation with more naturalistic stimuli/tasks would be a nice addition to this work. We have included a sentence in the discussion:

“All the effects observed in our experiments were consistent across near transfer conditions (rotation of patterns within the same feature space), and far transfer conditions (rotation of patterns within a different feature space, where features are drawn from the same modality). This shows the generality of spatial training for conceptual generalisation. We did not test transfer across modalities nor transfer in a more natural setting; we leave this for future studies.”

Further doubt on the underlying mechanism is cast by the possibility that the observed correlation between mapping task performance and the adoption of a 2D strategy may reflect general cognitive engagement rather than the spatial nature of the task. Similarly, the surprising finding that a significant number of participants benefited from spatial scaffolding without seeing spatial modalities may further raise questions about the interpretation of the scaffolding effect, pointing towards potential alternative interpretations, such as shifts in attention during learning induced by pre-training without changing underlying abstract conceptual representations.

The Reviewer is concerned about the fact that the spatial pre-training could benefit the participants by increasing global cognitive engagement rather than providing a scaffold for learning invariances. It is correct that the participants in the control group in Exp. 2c have poorer performances on average than participants that benefit from the spatial pre-training in Exp. 2a and 2b. The better performances of the participants in Exp. 2a and 2b could be due to either the spatial nature of the pre-training (as we claim) or a difference in general cognitive engagement. .

However, if we look closely at the results of Exp. 3, we can see that the general cognitive engagement hypothesis is not well supported by the data. Indeed, the participants in the control condition (Exp. 3c) have relatively similar performances than the other groups during training. Rather, the difference is in the strategy they use, as revealed by the transfer condition. The majority of them are using a 1D strategy, contrary to the participants that benefited from a spatial pre-training (Exp 3a and 3b). We have included a sentence in the results:

“Further, the results show that participants who did not experience spatial pre-training were still engaged in the task, but were not using the same strategy as the participants who experienced spatial pre-training (1D rather than 2D). Thus, the benefit of the spatial pre-training is not simply to increase the cognitive engagement of the participants. Rather, spatial pre-training provides a scaffold to learn rotation-invariant representation of auditory and visual concepts even when rotation is never explicitly shown during pre-training.”

Finally, Reviewer #1 had a related concern about a potential alternative explanation that involved a shift in attention. We reproduce our response here: we agree with the Reviewer that the “attention to dimensions” hypothesis is an interesting (and potentially concerning) alternative explanation. However, we believe that the results of our control experiments Exp. 2c and Exp. 3c are not compatible with this alternative explanation.

Indeed, in Exp. 2c, participants are pre-trained in the visual modality and then tested in the auditory modality. In the multimodal association task, participants have to associate the auditory stimuli and the visual stimuli: on each trial, they hear a sound and then have to click on the corresponding visual stimulus. It is necessary to pay attention to both auditory dimensions and both visual dimensions to perform well in the task. To give an example, the task might involve mapping the fundamental frequency and the amplitude modulation of the auditory stimulus to the colour and the shape of the visual stimulus, respectively. If participants pay attention to only one dimension, this would lead to a maximum of 25% accuracy on average (because they would be at chance on the other dimension, with four possible options). We observed that 30/50 participants reached an accuracy > 50% in the multimodal association task in Exp. 2c. This means that we know for sure that at least 60% of the participants actually paid attention to both dimensions of the stimuli. Nevertheless, there was a clear difference between participants that received a visual pre-training (Exp. 2c) and those who received a spatial pre-training (Exp. 2a) (frequency of 1D vs 2D models between conditions, BF > 100 in near transfer and far transfer). In fact, only 3/50 participants were best fit by a 2D model when vision was the pre-training modality compared to 29/50 when space was the pre-training modality. Thus, the benefit of the spatial pre-training cannot be due solely to a shift in attention toward both dimensions.

This effect was replicated in Exp. 3c. Similarly, 33/48 participants reached an accuracy > 50% in the multimodal association task in Exp. 3c, meaning that we know for sure that at least 68% of the participants actually paid attention to both dimensions of the stimuli. Again, there was a clear difference between participants who received a visual pre-training (frequency of 1D vs 2D models between conditions, Exp. 3c) and those who received a spatial pre-training (Exp. 3a) (BF > 100 in near transfer and far transfer).

Thus, we believe that the alternative explanation raised by the Reviewer is not supported by our data. We have added a paragraph in the discussion:

“One alternative explanation of this effect could be that the spatial pre-training encourages participants to attend to both dimensions of the non-spatial stimuli. By contrast, pretraining in the visual or auditory domains (where multiple dimensions of a stimulus may be relevant less often naturally) encourages them to attend to a single dimension. However, data from our control experiments Exp. 2c and Exp. 3c, are incompatible with this explanation. Around ~65% of the participants show a level of performance in the multimodal association task (>50%) which could only be achieved if they were attending to both dimensions (performance attending to a single dimension would yield 25% and chance performance is at 6.25%). This suggests that participants are attending to both dimensions even in the visual and auditory mapping case.”

Conclusions:

The authors successfully demonstrate that spatial training can enhance the ability to generalize in nonspatial domains, particularly in recognizing rotated sequences. The results for the most part support their conclusions, showing that spatial representations can act as a scaffold for learning more abstract conceptual invariances. However, the study leaves room for further investigation into whether the observed effects are unique to spatial cognition or could be replicated with other forms of well-established knowledge, as well as further clarifications of the underlying mechanisms.

Impact:

The study's findings are likely to have a valuable impact on cognitive science, particularly in understanding how abstract concepts are learned and generalized. The methods and data can be useful for further research, especially in exploring the relationship between spatial cognition and abstract conceptualization. The insights could also be valuable for AI research, particularly in improving models that involve abstract pattern recognition and conceptual generalization.

In summary, the paper contributes valuable insights into the role of spatial cognition in learning abstract concepts, though it invites further research to explore the boundaries and specifics of this scaffolding effect.

Reviewer #1 (Recommendations For The Authors):

Minor issues / typos:

P6: I think the example of the "signed" mapping here should be "e.g., ABAB maps to one category and BABA maps to another", rather than "ABBA maps to another" (since ABBA would always map to another category, whether the mapping is signed or unsigned).

Done.

P11: "Next, we asked whether pre-training and mapping were systematically associated with 2Dness...". I'd recommend changing to: "Next, we asked whether accuracy during pre-training and mapping were systematically associated with 2Dness...", just to clarify what the analyzed variables are.

Done.

P13, paragraph 1: "only if the features were themselves are physical spatial locations" either "were" or "are" should be removed.

Done.

P13, paragraph 1: should be "neural representations of space form a critical substrate" (not "for").

Done.

Reviewer #2 (Recommendations For The Authors):

The authors use in multiple places in the manuscript the phrases "learn invariances" (Abstract), "formation of invariances" (p. 2, para. 1), etc. It might be just me, but this feels a bit like 'sloppy' wording: we do not learn or form invariances, rather we learn or form representations or transformations by which we can perform tasks that require invariance over particular features or transformation of the input such as the case of object recognition and size- translation- or lighting-invariance. We do not form size invariance, we have representations of objects and/or size transformations allowing the recognition of objects of different sizes. The authors might change this way of referring to the phenomenon.

We respectfully disagree with this comment. An invariance occurs when neurons make the same response under different stimulation patterns. The objects or features to which a neuron responds is shaped by its inputs. Those inputs are in turn determined by experience-dependent plasticity. This process is often called “representation learning”. We think that our language here is consistent with this status quo view in the field.

Reviewer #3 (Recommendations For The Authors):

I understand that the objective of the present experiment is to study our ability to generalize abstract patterns of sensory data (concepts). In the introduction, the authors present examples like the concept of a "tree" (encompassing a family tree, an oak tree, and a taxonomic tree) and "ring" to illustrate the idea. However, I am sceptical as to whether these examples effectively represent the phenomena being studied. From my perspective, these different instances of "tree" do not seem to relate to the same abstract concept that is translated or rotated but rather appear to share only a linguistic label. For instance, the conceptual substance of a family tree is markedly different from that of an oak tree, lacking significant overlap in meaning or structure. Thus, to me, these examples do not demonstrate invariance to transformations such as rotations.

To elaborate further, typically, generalization involves recognizing the same object or concept through transformations. In the case of abstract concepts, this would imply a shared abstract representation rather than a mere linguistic category. While I understand the objective of the experiments and acknowledge their potential significance, I find myself wondering about the real-world applicability and relevance of such generalizations in everyday cognitive functioning. This, in turn, casts some doubt on the broader relevance of the study's results. A more fitting example, or an explanation that addresses my concerns about the suitability of the current examples, would be beneficial to further clarify the study's intent and scope.

Response in the public review.

Relatedly, the manuscript could benefit from greater clarity in defining key concepts and elucidating the proposed mechanism behind the observed effects. Is it plausible that the changes observed are primarily due to shifts in attention induced by the spatial pre-training, rather than a change in the process of learning abstract conceptual invariances (i.e., modifications to the abstract representations themselves)? While the authors conclude that spatial pre-training acts as a scaffold for enhancing the learning of conceptual invariances, it raises the question: does this imply participants simply became more focused on spatial relationships during learning, or might this shift in attention represent a distinct strategy, and an alternative explanation? A more precise definition of these concepts and a clearer explanation of the authors' perspective on the mechanism underlying these effects would reduce any ambiguity in this regard.

Response in the public review.

I am wondering whether the effectiveness of spatial representations in generalizing abstract concepts stems from their special nature or simply because they are a familiar 2D space for participants. It is well-established that memory benefits from linking items to familiar locations, a technique used in memory training (method of loci). This raises the question: Are we observing a similar effect here, where spatial dimensions are the only tested familiar 2D spaces, while the other 2 spaces are simply unfamiliar, as also suggested by the lower performance during training (Fig.2)? Would the results be replicable with another well-learned, robustly encoded domain, such as auditory dimensions for professional musicians, or is there something inherently unique about spatial representations that aids in bootstrapping abstract representations?

On the other side of the same coin, are spatial representations qualitatively different, or simply more efficient because they are learned more quickly and readily? This leads to the consideration that if visual pre-training and visual-to-auditory mapping were continued until a similar proficiency level as in spatial training is achieved, we might observe comparable performance in aiding generalization. Thus, the conclusion that spatial representations are a special scaffold for abstract concepts may not be exclusively due to their inherent spatial nature, but rather to the general characteristic of well-established representations. This hypothesis could be further explored by either identifying alternative 2D representations that are equally well-learned or by extending training in visual or auditory representations before proceeding with the mapping task. At the very least I believe this potential explanation should be explored in the discussion section.

Response in the public review.

I had some difficulty in following an important section of the introduction: "... whether participants can learn rotationally invariant concepts in nonspatial domains, i.e., those that are defined by sequences of visual and auditory features (rather than by locations in physical space, defined in Cartesian or polar coordinates) is not known." This was initially puzzling to me as the paragraph preceding it mentions: "There is already good evidence that nonspatial concepts are represented in a translation invariant format." While I now understand that the essential distinction here is between translation and rotation, this was not immediately apparent upon first reading. This crucial distinction, especially in the context of conceptual spaces, was not clearly established before this point in the manuscript. For better clarity, it would be beneficial to explicitly contrast and define translation versus rotation in this particular section and stress that the present study concerns rotations in abstract spaces.

Done.

The multi-modal association is crucial for the study, however to my knowledge, it is not depicted or well explained in the main text or figures (Results section). In my opinion, the details of this task should be explained and illustrated before the details of the associated results are discussed.

We have included an illustration of a multimodal association trial in Fig. S3B.

The observed correlation between the mapping task performance and the adoption of a 2D strategy is logical. However, this correlation might not exclusively indicate the proposed underlying mechanism of spatial scaffolding. Could it also be reflective of more general factors like overall performance, attention levels, or the effort exerted by participants? This alternative explanation suggests that the correlation might arise from broader cognitive engagement rather than specifically from the spatial nature of the task. Addressing this possibility could strengthen the argument for the unique role of spatial representations in learning abstract concepts, or at least this alternative interpretation should be mentioned.

Response in the public review.

To me, the finding that ~30% of participants benefited from the spatial scaffolding effect for example in the auditory condition merely through exposure to the mapping (Fig 4D), without needing to see the quadruplets in the spatial modality, was somewhat surprising. This is particularly noteworthy considering that only ~60% of participants adopted the 2D strategy with exposure to rotated contingencies in Experiment 3 (Fig 3D). How do the authors interpret this outcome? It would be interesting to understand their perspective on why such a significant effect emerged from mere exposure to the mapping task.

I appreciate the clarity Fig.1 provides in explaining a challenging experimental setup. Is it possible to provide example trials, including an illustration that shows which rotations produce the trail and an intuitive explanation that response maps onto the 1D vs 2D strategies respectively, to aid the reader in better understanding this core manipulation?

I like that the authors provide transparency by depicting individual subject's data points in their results figures (e.g. Figs. 2 B, F, J). However, with an n=~50 per condition, it becomes difficult to intuit the distribution, especially for conditions with higher variance (e.g., Auditory). The figures might be more easily interpretable with alternative methods of displaying variances, such as violin plots per data point, conventional error shading using 95%CIs, etc.

Why are the authors not reporting exact BFs in the results sections at least for the most important contrasts?

While I understand why the authors report the frequencies for the best model fits, this may become difficult to interpret in some sections, given the large number of reported values. Alternatives or additional summary statistics supporting inference could be beneficial.

As the Reviewer states, there are a large number of figures that we can report in this study. We have chosen to keep this number at a minimum to be as clear as possible. To illustrate the distribution of individual data points, we have opted to display only the group's mean and standard error (the standard errors are included, but the substantial number of participants per condition provides precise estimates, resulting in error bars that can be smaller than the mean point). This decision stems from our concern that including additional details could lead to a cluttered representation with unnecessary complexity. Finally, we report what we believe to be the critical BFs for the comprehension of the reader in the main text, and choose a cutoff of 100 when BFs are high (corresponding to the label “decisive” evidence, some BFs are larger than 1012). All the exact BFs are in the supplementary for the interested readers.

Version 2

Abstract

Abstract

Learning invariances allows us to generalise. In the visual modality, invariant representations allow us to recognise objects despite translations or rotations in physical space. However, how we learn the invariances that allow us to generalise abstract patterns of sensory data (“concepts”) is a longstanding puzzle. Here, we study how humans generalise relational patterns in stimulation sequences that are defined by either transitions on a nonspatial two-dimensional feature manifold, or by transitions in physical space. We measure rotational generalisation, that is the ability to recognise concepts even when their corresponding transition vectors are rotated. We find that humans naturally generalise to rotated exemplars when stimuli are defined in physical space, but not when they are defined as positions on a nonspatial feature manifold. However, if participants are first pre-trained to map auditory or visual features to spatial locations, then rotational generalisation becomes possible even in nonspatial domains. These results imply that space acts as a scaffold for learning more abstract conceptual invariances.

Introduction

To recognise objects and events in the natural world, humans form mental representations that are invariant to transformation. The existence of invariant representations allows entities to be recognised and categorised despite changes in their surface properties, which is called “generalisation”. The formation of invariances has been most extensively studied in the case of visual object recognition. For example, we have no trouble recognising a teapot that is moved to a new location (translated), tipped on its side (rotated) or viewed from afar (rescaled). How we do so has provoked diverse theories based on assembly from geometric primitives [1,2], associative learning [3,4], and function approximation in deep networks [5].

A core problem in cognitive science, however, is how we form invariances over entities that are defined by more abstract relational properties. Here, we use the term “concepts” to refer to objects or events that are defined by shared relations among features that may unfold in space, or time, or both, in any modality [6]. For example, the concept of a “tree” implies an entity whose structure is defined by a nested hierarchy, whether this is a physical object whose parts are arranged in space (such as an oak tree in a forest) or a more abstract data structure (such as a family tree or taxonomic tree). The concept of a “ring” implies an entity whose features are arranged cyclically, whether a physical ring (worn on the finger), the (circular) temporal pattern of tones in a peal of bells, or the periodicity in the passage of the seasons [7]. Despite great changes in the surface properties of oak trees, family trees and taxonomic trees, humans perceive them as different instances of a more abstract concept defined by the same relational structure. The human ability to readily form invariances over abstract concepts remains a puzzle for both cognitive and neural scientists hoping to understand neural computations, and a challenge for AI researchers wishing to build intelligent agents.

One prominent theory argues that we learn invariant concepts because of the way the brain represents physical space [8–12]. This argument states that neurons coding for positions in either egocentric (viewer-centred) or allocentric (world-centred) space can be recycled to represent locations in more abstract spaces, defined by continuous variation in features (e.g., red to blue, quiet to loud). This theory is backed up by proof-of-concept computational simulations [13], and by findings that brain regions thought to be critical for spatial cognition in mammals (such as the hippocampal-entorhinal complex and parietal cortex) exhibit neural codes that are invariant to relational transformations of nonspatial stimuli [14–18]. However, whilst promising, this theory lacks direct empirical evidence. Here, we set out to provide a strong test of the idea that learning about physical space scaffolds conceptual generalisation. Our focus is on the ability to generalise knowledge about the relations among items in a sequence as they are translated or rotated through both spatial and nonspatial domains.

In the four studies described here, participants made category judgments about a sequence of four successive stimuli in either auditory, visual or spatial modalities. In auditory and visual modalities, the stimulus was drawn from a two-dimensional feature manifold (e.g., a bivariate “space” defined by colour and shape in the visual modality or pitch and timbre in the auditory modality). In the spatial modality, each stimulus was a position in physical space (e.g., an x and y coordinates). Concepts were defined by a common pattern of transitions through either feature space or physical space. Our research question concerned the conditions under which concepts could be recognised, even if their corresponding transition vectors had been translated or rotated. We studied generalisation of transition vectors both within the same feature space and to new feature spaces in the same modality.

Our studies measure the tendency to generalise both by translation and rotation. Conceptual translation occurs when feature values are shifted in either dimension, but with no change in their relational pattern. There is already good evidence that nonspatial concepts are represented in a translation invariant format. For example, in the auditory domain, we can recognise “auditory objects” that are translated in feature space (e.g., pitch and timbre). This occurs when we understand the same sentence from different speakers, or identify the same melody played with different musical instruments [19,20]. However, much less is known about the learning of rotational invariances for abstract concepts. In physical space, we readily learn rotation-invariant object representations (allowing us to recognise an upside down teapot), and the computational mechanisms by which we do so have been a major fulcrum of debate in the vision sciences [3,4]. But whether participants can learn rotationally invariant concepts in nonspatial domains, i.e. those that are defined by sequences of visual and auditory features (rather than by locations in physical space, defined in Cartesian or polar coordinates) is not known. In the current study, we first test this, and find that naively, they cannot. Next, turning to our main hypothesis, we then ask if first teaching participants to map nonspatial features to spatial locations (providing a spatial scaffold) allows the learning of rotational invariances, even in nonspatial modalities. We find that it does. This shows that a form of generalisation that is not usually possible for humans becomes possible when their understanding of the concept is “scaffolded” by first learning a corresponding spatial representation. This thus supports the theory that abstract concept learning is linked to our understanding of physical space.

Results

On each trial, participants were presented with a sequence of four auditory, visual or spatial stimuli (a quadruplet) drawn from one of 16 points on a continuously varying 2D (4 x 4) feature manifold. In the visual (auditory) modality, this manifold was respectively defined by two orthogonal and continuously varying visual (auditory) features. In the spatial domain, the 2D feature manifold was defined by positions in physical space, in either Cartesian or polar coordinates (see Fig. 1D, Fig. S1-S2). Each quadruplet was constructed by first sampling a random point on the 2D feature manifold, and then iteratively choosing three further adjacent feature locations to make a sequence of four stimuli. In each experiment, there were three categories (Fig. 1C). Each category was initially defined by a canonical set of transition vectors, which specified the three successive steps on the feature manifold (defining the positions from which stimuli in the quadruplet were sampled). Thus, for example, one set of transition vectors might be defined by compass directions {NE, W, SE}. This would mean that after an initial stimulus was sampled, the second stimulus in the sequence would be the one NE in feature space, and the third W of that, and the fourth SE of that). We define rotational generalisation as the ability to recognise regularities in the sequence transition vectors that are independent of both translation and rotation. Thus, just as an upside-down teapot can still be recognised by the relative spatial relations among its handle, body, and spout, in Exp. 1 we asked whether concepts can be recognised when their associated transition vectors are rotated (e.g., vector sequence {NE, W, SE} on the feature manifold becomes {NW, S, NE} after 90° rotation). Note that in our study, quadruplets are also randomly translated on the manifold by virtue of the variable initial feature selection between trials. We thus make the basic assumption that rotational generalisation also involves translation invariance.

Paradigm.

A. Objects can be perceived as similar despite changes in shape, orientation and features. B. Similarly, can sequences of features be perceived as similar despite great changes in the features that composed them? C. In our experiments, participants had to learn the category associated with quadruplets composed of four stimuli drawn from a 2D feature manifold. To do so, they could use one of two major strategies: tracking changes in a single dimension (1D) or two dimensions (2D). D. Illustration of the feature manifold and transition vectors in the visual modality. Each transition vector defined transitions between cardinally or diagonally adjacent features on a 2D manifold (here, given by colour and spikiness for training and near transfer in the left and middle panels, and by transparency and squareness in the far transfer condition). The transition vectors for each category are shown in their canonical (0°) rotation as blue, orange and green arrows superimposed on each feature manifold. The vectors are shown rotated by 90° in the near and far transfer conditions. Next to each feature manifold is a matrix showing the expected mapping (filled squares) from feature vectors to categories for a participant using the 1D strategies (top and middle) and 2D strategy (bottom). Note that the 2D strategy always leads to effective generalisation (rotated exemplars are mapped on their corresponding categories) even in transfer, as indicated by filled squares on the matrix diagonal, whereas a 1D strategy leads to a different pattern. We use the symbols ∅, ↶ and ★ to denote canonical (0°), 90° rotation and far transfer conditions respectively.

Our basic procedure was as follows. During training (120 trials), participants first learned to assign canonical (0° rotation) quadruplets to one of three categories using a button press, receiving fully informative feedback after each response (see Fig. S3). Then, during test, participants performed a further 210 trials, half of which were identical to training (with feedback) while the other half were transfer trials involving categorisation of quadruplets whose feature transition vectors were rotated by 90°, 180° or 270°. These novel quadruplets were either sampled from the same 2D feature manifold (e.g., colour and spikiness in the visual case; near transfer condition) or a new 2D feature manifold from the same modality (e.g., transparency and squareness; far transfer condition; see Fig. 1D, Fig. S2). Transfer trials received no feedback, allowing us to infer what knowledge was being generalised between training and transfer. Experiments 1-3 were pre-registered at https://osf.io/z9572/registrations.

Concepts defined by spatial locations, but not auditory or visual features, are rotation-invariant

In Exp. 1, we recruited three cohorts of online participants (N = 50 each, see Fig. 2) to perform the task in the auditory, visual and spatial modalities. These conditions differed only in how the feature manifold was defined: e.g., fundamental frequency and modulation frequency for auditory features (Fig. 2A); e.g., spikiness and colour for visual features (Fig. 2E); e.g., horizontal and vertical position for spatial locations (Fig. 2I). Accuracy on training trials for each modality is shown in Fig. 2B, F, J. Participants learned the task well in all three conditions (but better in the spatial modality: intercept β = 2.90 ± 0.19, slope associated with the auditory modality β = -1.91 ± 0.27, p < 0.001, slope associated with the visual modality, β = -1.57 ± 0.26, p < 0.001, mixed logistic regression on the probability of a correct response with participants as random effect). However, our main question was how participants would generalise learning to novel, rotated exemplars of the same concept.

Experiment 1: Quadruplets composed of spatial locations, but not auditory or visual features, are associated with a 2D strategy.

A. In the auditory modality (N = 50), the 2D feature manifold was defined by auditory features. Here, we illustrate with a schematic in which the feature manifold is defined by fundamental frequency and frequency modulation. B. Accuracy on training trials involving the canonical transition vectors (0° rotation, denoted by the symbol ∅). The bold curve and dots represent the group average; lighter curves are individual participants. During training, random models (R) are at chance (33% accuracy; lower dashed line), while idealised 1D and 2D models are at ceiling (100% accuracy; upper dashed lines). C. Model fits to the near transfer responses. The bar plot shows model frequencies in the population and black dots are Bayesian point estimates of the model frequency. The matrices show cross-validated model predictions in the same format at Fig. 1D, except with the degree of fading (light to dark colour) signalling average participant responses in each case. The matrices read as follows: the top row depicts the average behaviour of participants who were best fit by the 1Diu model (using a held out set of responses). The 1Diu model predictions are shown in black for three rotations of the quadruplets (90°, 180°, 270° denoted by the symbols ↶, ↺ and ↷ respectively). The average response matrices of the participants using held-out responses are shown in red. The scatter plot below shows individual likelihoods for the 1D and 2D models (normalised by the R model). Each dot is an individual participant, coded by whether they are best fit by either 1D or 2D models (colour) or the R model (grey). Dashed lines distinguish zones of relative likelihood where participants are best fit by R (bottom left square), 1D (rightmost quadrilateral) or 2D (upper quadrilateral) models. D. Model fits to the far transfer responses, using the same conventions as C. Note that, at transfer, more 1D models are possible because of the differing ways that participants could map between the i and j axes and the ★i and ★j axes during training and far transfer. E. In the visual modality (N = 52), the 2D feature manifold was defined by visual attributes; here we use colour and spikiness as an illustration. F-G-H depict data from the visual modality using conventions from B-C-D. I. In the spatial modality (N = 51), the 2D feature manifold was defined by the spatial location of a star (here a red dot is used for illustration). J-K-L depict data from the spatial modality using conventions from B-C-D.

To test this, we fit a family of quantitative models jointly to the training and transfer trials. To understand the logic of this modelling exercise, it is necessary to consider the alternative strategies that participants may have learned during training. Whereas rotation requires participants to represent both dimensions of the feature manifold (a rotation of 90° is only discernible in 2D), a viable alternative strategy during training is to base categorical decisions on a single feature (e.g., either spikiness or colour but not both). Each quadruplet consists of four adjacent feature locations forming a square on the feature manifold (Fig. 2D) and thus the stimulation sequence comprises two features from each dimension. Thus, for example, if a participant attended only to spikiness, the four stimuli in a quadruplet would be represented as a feature pattern over spikiness levels (such as ABAB or ABBA, where A = more spiky, B = less spiky). During training, participants could learn to map these patterns onto categories, either in a signed fashion (e.g., ABAB maps to one category and BABA to another), or an unsigned fashion (ABAB and BABA both map to the same category). These strategies would lead to perfect performance during training, but would prevent the learning of rotational invariances. We built models that implemented these one-dimensional strategies, which we call 1Ds and 1Du respectively, and compared them to models that used both dimensions for categorisation (the 2D model) or were simply responding randomly (R models). Each of these models predicts a unique pattern of generalisation (Fig. S4-S5) and only the 2D model predicts that participants will assign rotated objects to the same category as their unrotated counterparts (rotational generalisation). Thus, the principal metrics we report in this study are the fraction of (non-random) subjects classified as 1D vs. 2D on transfer trials, which is a signature of whether the experimental conditions permitted the learning of rotational invariances for quadruplets. We report both fractions of participants (X/X best fit by each model) and Bayes Factors reflecting the relative likelihood of 1D vs. 2D models between conditions.

Consistent with our first pre-registered prediction, Exp. 1 revealed a striking dissociation in rotational generalisation between modalities. For near transfer, all non-random participants in the auditory and visual modality (26/26 and 38/38) learned a 1Du strategy, whereas the vast majority in the spatial modality (35/41) were best fit by a 2D strategy. Bayesian group model comparison confirmed that the frequency of 1D vs 2D models among non-random participants was similar between the auditory and visual modalities (Bayes Factor [BF] = 0.1, “negative” evidence for a difference) but different between the auditory and spatial modalities (BF > 100, “decisive” evidence for a difference) and the visual and spatial modalities (BF > 100; see tables S6-S9 for full results). This implies that the use of a 1D strategy (implying no rotational generalisation) was much more likely than a 2D strategy (implying rotational generalisation) when the manifold was defined by visual or auditory features (e.g., colour and shape or pitch and timbre), but the converse was true when the feature manifold was defined by coordinates in physical space (e.g., horizontal and vertical position).

For far transfer, the results were very similar. In the auditory modality, all non-random participants (26/26) were again best fit by a 1Du strategy, and in the visual modality, most (30/35) were fit by a 1Du strategy, 5/35 by a 1Ds strategy, and none by a 2D strategy (difference between auditory and visual, BF = 0.1). By contrast, in the spatial modality, where far transfer involved remapping from cardinal to polar coordinates or vice versa, almost all non-random participants (29/31) were again best fit by a 2D strategy (both BF > 100 comparing with the auditory and visual modalities, “decisive” evidence for a difference). Behaviour in each modality of Exp. 1 is illustrated in Fig. 2, where we display category assignments under each rotation for participants allocated to distinct model classes on the basis of held out data. Together, these data show definitively that, when categories were characterised by temporal patterns in spatial location (e.g., where transitions in physical space were aligned with those on the feature manifold), participants learned to represent the 2D structure of the concept, and generalised readily to rotated (as well as translated) exemplars. However, when concepts were defined by patterns of nonspatial auditory or visual features, participants learned mappings to each category by relying on a single feature dimension and thus failed to form rotational invariant representations.

Spatial pre-training provides a scaffold for rotational generalisation in the auditory and visual modalities

Exp. 1 shows that rotational generalisation succeeds for spatial concepts but fails for nonspatial concepts. Next, in Exp. 2 we tested our main prediction: that space can be used as a scaffold for rotational generalisation of nonspatial concepts. We recruited three new cohorts of participants (N = 50 each, see Fig. 3) to perform a multi-phase task that unfolded over two successive days. On day 1, participants received 60 pre-training trials in the pre-training modality. These trials matched training trials in Exp. 1 for the corresponding modality (spatial or visual) except that they comprised both canonical (0°) and 90° rotated quadruplets, but not those rotated by 180° or 270° (we included examples of rotated quadruplets in the training set to encourage rotational generalisation, but as shown in Exp. 3, results do not depend on this choice). Subsequently, participants performed 288 trials of a multimodal association task, in which they learned the association between each of the 16 stimuli in the pre-training modality and their corresponding stimulus in a different testing modality, where the corresponding stimulus occupied an equivalent position on the 2D feature manifold (we call this the “mapping task”). The goal of this task was to teach participants correspondences between either spatial and visual, spatial and auditory, or visual and auditory feature manifolds. Then on day 2, after some refresher pre-training and mapping trials, participants performed the same task as in Exp. 1 in the testing modality, again with the exception that training trials also included 90° rotated quadruplets.

Experiment 2: Spatial pre-training triggers the use of a 2D strategy for quadruplets composed of auditory or visual features.

A. Performance on the spatial pre-training task in Exp. 2a (N = 51). For this and all plots below, the bold line is the group average and lighter lines are individual participants. Dashed lines show expected average performance under the corresponding models (labelled). B. Performance in the spatial to auditory mapping task. Chance performance is at 1/16 (dashed line). C. Performance on training trials in the auditory modality for canonical (0°) and 90° rotated quadruplets. During training, random models (R) are at chance (33% accuracy), almost all 1D models are at 50% accuracy, and 1Dis and 2D models are at ceiling (100% accuracy). D. Model fits to the near transfer responses, using the same conventions as Fig. 2. Model predictions are shown in black for two rotations of the quadruplets (180° and 270° denoted by the symbols ↺ and ↷ respectively). E. Model fits to the far transfer responses. F-G-H-I-J read as A-B-C-D-E for Exp. 2b (N = 51), where participants performed a spatial pre-training and a spatial to visual mapping task prior to performing the visual version of the main task. K-L-M-N-O read as A-B-C-D-E for Exp. 2c (N = 50), where participants performed a visual pre-training and a visual to auditory mapping task prior to performing the auditory version of the main task.

Our pre-registered prediction for Exp. 2 was that when (Cartesian) physical space was the pre-training modality, participants would now (in contrast to Exp. 1) learn using a predominantly 2D strategy in both auditory (Exp. 2a) and visual (Exp. 2b) testing modalities. In other words, by learning the association between auditory or visual features and a corresponding spatial location, concepts composed of exclusively nonspatial features could now be generalised over rotations in a way not exhibited by a single participant in Exp. 1a or Exp. 1b. By way of control, however, we predicted that when the pre-training involved (nonspatial) visual features, no such benefit would occur, and participants would fail to show rotation invariance.

This is exactly what we found, for both near and far transfer. In the near transfer condition, with spatial pre-training, 29/37 non-random participants were best fit by a 2D strategy when audition was the testing modality, and 36/40 when vision was the testing modality. By contrast however, participants who underwent visual (rather than spatial) pre-training failed to show a benefit when audition was the testing modality. In fact, most (37/50) were best fit by the random model (see below), with 6/13 non-random participants favouring a 1Ds strategy. We once again calculated Bayes Factors at the group level to assess the reliability of these results. We found that BFs exceeded 100, providing “decisive” evidence that the 2D model was more favoured among the groups with spatial pre-training than that without. Similarly, in the far transfer condition, spatial pre-training allowed 29/38 participants in the auditory modality and 13/16 participants in the visual modality to successfully generalise via a 2D strategy. This was not the case for participants who experienced visual pre-training (again, frequency of 1D vs 2D models between conditions: BF > 100, “decisive”). In other words, spatial pre-training provided an effective scaffold that allowed participants to learn auditory and visual objects in a 2D representational format that permitted generalisation to novel rotated exemplars.

Participants in Exp. 2c performed poorly during training and were more likely to be fit by the random models during transfer than those who performed the same auditory task in Exp. 1a. Indeed, we computed the Bayes Factor quantifying the relative likelihood of the random (R models) vs all other models (1D and 2D models) and found “substantial” evidence in favour of a difference between groups both in near transfer (BF = 7.4) and in far transfer (BF = 2.7). This might seem curious, because Exp. 2 participants had access to more diverse training (on both 0° and 90° quadruplets) as well as the supplementary visual pre-training. Why did Exp. 2c participants struggle with the task? In fact, this phenomenon makes sense, because training with feedback on both 0° and 90° quadruplets effectively invalidates a 1D strategy, because there no longer exists a unique mapping between categories and features in either of the two feature dimensions (note that training performance in Exp. 2c plateaus close to 50%). This lack of a viable 1D strategy during training obliges participants to use a 2D strategy where possible. Because this is only possible with spatial pre-training, in Exp. 2c they revert to random. Whilst this explains what we observed in Exp. 2, it also allows a further prediction: if we remove the 90° rotated quadruplets from pre-training, then participants in the spatial pre-training modality should be somewhat less prone to use a 2D strategy (because 1D is available) whereas participants who undergo visual pre-training should show more 1D behaviour at the expense of the random model. In Exp. 3, we tested and confirmed this prediction.

Exp. 3 involved three new cohorts (N = 50 each, see Fig. 4) and was identical to Exp. 2, except that now pre-training trials consisted exclusively of canonical (0°) quadruplets (although 90° quadruplets were still present when the testing modality was trained on day 2). As predicted, non-random participants who enjoyed spatial pre-training were still prone to use a 2D strategy when audition was the testing modality (16/32 for near transfer and 17/31 for far transfer) as well as when vision was the testing modality (22/30 and 13/19), replicating the findings of Exp. 2. However, compared to Exp. 2, overall more participants relied on 1D strategies. In the auditory modality in Exp. 3a, 16/32 were best fit by a 1D model in the near transfer condition (9/32 1Du and 7/32 1Ds) and 14/31 in the far transfer condition (10/31 1Du and 4/31 1Ds). At the group level, the Bayes Factor confirmed that participants were more likely to be fit by a 1D model in Exp. 3a than Exp. 2a in the auditory modality (frequency of 1D vs 2D models between Exp. 2a and Exp. 3a, near transfer BF = 3.4 “substantial” evidence, far transfer BF = 1.5 “weak” evidence). Similarly, in the visual modality in Exp. 3b, 8/30 were best fit by a 1D model in the near transfer condition (5/30 1Du and 3/30 1Ds) and 6/19 in the far transfer condition (6/19 1Du) (again, frequency of 1D vs 2D models between Exp. 2b and Exp. 3b, near transfer BF = 5.9 “substantial” evidence, far transfer BF = 2.2 “weak” evidence). By contrast, participants who underwent nonspatial (visual) pre-training did not use a 2D strategy (1/30) but rather preferred 1D strategies in both near transfer (13/30 1Du and 16/30 1Ds) and far transfer conditions (10/28 1Du and 17/28 1Ds). Comparing these results with the frequency of 1D vs 2D models in conditions with spatial pre-training (Exp. 3a and 3b), we found that all BFs exceeded 100, providing “decisive” evidence that the 2D model was more favoured among the groups with spatial pre-training than that without.

Experiment 3: Exposure to the canonical events (0°) during spatial pre-training is sufficient to trigger the use of a 2D strategy.

All panels use the same conventions as Fig. 3. A-B-C-D-E. Contrary to Exp. 2a, in Exp. 3a (N = 50), participants were exposed only to canonical events (0°) during the spatial pre-training. F-G-H-I-J. Exp. 3b (N = 51). K-L-M-N-O. Exp. 3c (N = 50).

Thus, these results show that training exclusively on canonical (0°) quadruplets facilitates a 1D strategy, which is expressed more readily than in Exp. 2; but that the 2D strategy is still more likely for participants who underwent spatial pre-training. Further, the results show that participants who did not experience spatial pre-training were still engaged in the task, but were not using the same strategy as the participants who experienced spatial pre-training (1D rather than 2D). Thus, the benefit of the spatial pre-training is not simply to increase the cognitive engagement of the participants. Rather, spatial pre-training provides a scaffold to learn rotation-invariant representation of auditory and visual concepts even when rotation is never explicitly shown during pre-training. Furthermore, participants are sensitive to the available strategies during pre-training, and use the 1D strategy when possible if they have not learned to associate features with space.

Spatial mapping performance predicts rotational generalisation for nonspatial modalities

Next, we used our data from Exp. 2 and Exp. 3 to study how performance on each phase of our task predicted rotational generalisation in the testing phase (see Fig. 5). For each participant, we created an index of rotation generalisation (2Dness) as the difference in log-likelihood between the best 1D model and the 2D model during near transfer. We found that 2Dness was powerfully predicted by training accuracy (Pearson correlation between 2Dness and training accuracy [r2Dness,TRAINING] = 0.80, p < 0.001) in both Exp. 2 (r2Dness,TRAINING = 0.83, p < 0.001) and Exp. 3 (r2Dness,TRAINING = 0.78, p < 0.001). The fact that training performance is highly correlated with 2Dness implies that participants who solved the training task formed representations that were generalisable in 2D; in other words, very few participants overfit to the training set. Accordingly, participants were poorly captured by an additional model (the R’ model; 4/152 in Exp. 2, 4/151 in Exp. 3), that has perfect performance during training but responds randomly during transfer. Next, we asked whether accuracy during pre-training and mapping were systematically associated with 2Dness, and assessed their relative importance using partial correlations. Pre-training did explain unique variance in 2Dness after accounting for mapping (correlation between pre-training and 2Dness after partialling out mapping [r2Dness,PRETRAINING - MAPPING] = 0.17, p < 0.01) and vice versa (r2Dness,MAPPING - PRETRAINING = 0.27, p < 0.001). However, 2Dness was better predicted by mapping than by pre-training in Exp. 2a (r2Dness,MAPPING - PRETRAINING = 0.42, p < 0.005 and r2Dness,PRETRAINING - MAPPING = 0.30, p < 0.05), Exp. 2b (r2Dness,MAPPING - PRETRAINING = 0.41, p < 0.005 and r2Dness,PRETRAINING - MAPPING = 0.04, p = 0.81), Exp. 3a (r2Dness,MAPPING - PRETRAINING = 0.32, p < 0.05 and r2Dness,PRETRAINING - MAPPING = 0.06, p = 0.70) and Exp. 3b (r2Dness,MAPPING - PRETRAINING = 0.46, p < 0.001 and r2Dness,PRETRAINING - MAPPING = 0.33, p < 0.05). The strong correlations between the mapping task performances and 2Dness suggest that learning the association between nonspatial and spatial features is the critical step that allows rotational generalisation.

Correlation analysis.

Scatter plots of 2Dness (the difference in likelihood between the best 1D model and the 2D model in the training modality in near transfer) and A. pre-training accuracy, B. mapping accuracy and C. training accuracy in the testing modality in Exp. 2. Dots are individual participants in Exp. 2a (red), 2b (blue) and 2c (brown). D-E-F read as A-B-C for Exp. 3.

We tested and confirmed this prediction in Exp. 4 (see Fig. S7) which repeated Exp. 3 except that spatial pre-training was replaced with a duration-matched filler task (in which the category is defined by the number of stationary blue stars in a sequence). Without spatial pre-training, a sizeable proportion of participants still learned a 2D strategy in both the auditory (9/30 in near transfer, 9/28 in far transfer) and visual (12/19 and 7/19) modality, although the majority relied on a 1D strategy (auditory modality: 4/30 1Du and 17/30 1Ds for near transfer, 4/38 1Du and 15/28 1Ds for far transfer; visual modality: 5/19 1Du and 6/19 1Ds for near transfer, 5/14 1Du and 2/14 1Ds for far transfer). In the auditory modality (Exp. 4a), this can be compared with Exp. 2c, where almost all participants were using a random strategy (frequency of R vs 1D/2D models, BF = 34.0, “strong” evidence), and with Exp. 3c where almost no participants were using a 2D strategy (frequency of 1D vs 2D models, BF = 8.6, “substantial” evidence). Thus, for ~20% participants, the mere exposure to the mapping was sufficient to benefit from the spatial scaffolding effect and actually seeing the quadruplets in the spatial modality was not necessary for them.

Discussion

We studied the conditions under which participants learn rotation- and translation-invariant representations of abstract concepts. We found that participants can generalise conceptual knowledge to novel sequences (quadruplets) defined by rotations of stimulus feature transition vectors, but only if the features were themselves physical spatial locations (e.g., x, y position; Exp. 1) or if nonspatial attributes had previously been mapped to a physical spatial location in a pre-training task (Exp. 2-4). Thus, an explicit representation of physical space is a “scaffold” that permits objects to be learned in a rotation invariant fashion, and thus allows rotational generalisation. This supports the idea that neural representations of space form a critical substrate for learning abstractions in nonspatial domains [8–10,21].

It is well known that humans learn rotational invariances for visual objects, whose features are organised in physical space. For example, an upside down teapot can be recognised by the relative position of handle, lid and spout. This case mimics our spatial modality condition, where each concept was a pattern of locations in physical space. It is thus perhaps unsurprising that rotational generalisation is possible in this condition. However, we found it striking that participants generalised in such different ways when the features in question were drawn from a nonspatial manifold, in either the visual or auditory domain. In these conditions, participants seemed to have no trouble recognising patterns that were consistently translated in feature space. This is consistent with previous studies that have shown that we can understand language in different accents, or name a familiar tune played at an atypical speed or pitch [22]. However, they did so via a representation that focused on just one of the two possible dimensions, and thus did not permit rotational generalisation. There was thus a clear dissociation between human ability to generalise patterns in physical space and a more abstract feature space.

Next, we showed that spatial pre-training allowed rotational generalisation even for sequences composed of nonspatial features. This implies that the neural representation of space may serve as a “scaffold”, allowing people to visualise and manipulate nonspatial concepts. One alternative explanation of this effect could be that the spatial pre-training encourages participants to attend to both dimensions of the non-spatial stimuli. By contrast, pretraining in the visual or auditory domains (where multiple dimensions of a stimulus may be relevant less often naturally) encourages them to attend to a single dimension. However, data from our control experiments Exp. 2c and Exp. 3c, are incompatible with this explanation. Around ~65% of the participants show a level of performance in the multimodal association task (>50%) which could only be achieved if they were attending to both dimensions (performance attending to a single dimension would yield 25% and chance performance is at 6.25%). This suggests that participants are attending to both dimensions even in the visual and auditory mapping case. Rather, whilst we are not aware of previous studies that have tested spatial scaffolding in the way described here, our findings are consistent with the more general idea that space is represented in an overlapping fashion with nonspatial information, such as time or number [23]. For example, sequences with regular spatial geometry are learned more readily than those composed of arbitrary patterns [24]. Our findings also cohere with evidence that visuospatial skills are correlated with a variety of academic competences, especially in STEM subjects such as maths and engineering [25], and that spatial training interventions (such as teaching mental rotation) in educational settings can improve nonspatial abilities, such as calculus grades [26].

The idea that spatial representations form a generalised substrate for cognition – including for coding temporal structure – draws on a long tradition in philosophy [27], cognitive science [11] and neuroscience [8–10]. The precise substrate for this effect is unclear, but it seems likely that neural assemblies activated by physical locations in space (for example, in parietal or medial temporal lobe areas) are recycled for representing nonspatial patterns in data. We acknowledge that our study does not provide a mechanistic model of the spatial scaffolding effect but rather delineate which aspects of the training are necessary for generalisation to happen. In our study, thus, the mapping task facilitates this recycling by teaching participants a point-to-point mapping between nonspatial feature combinations and locations in physical space. Indeed, our correlation analysis and Exp. 4 suggested that successfully learning mappings between spatial and nonspatial features was the strongest determinant of rotational generalisation. This mapping task was presented in an egocentric frame of reference defined by the x, y coordinates of the screen. Explicit representations of location in egocentric space in the primate are found in dorsal stream structures such as the posterior parietal cortex [28]. Current deep networks – which successfully categorise lone objects in a natural image but often fail on tests of relational reasoning or scene understanding – may be hampered by their failure to represent space explicitly in this way [10].

All the effects observed in our experiments were consistent across near transfer conditions (rotation of patterns within the same feature space), and far transfer conditions (rotation of patterns within a different feature space, where features are drawn from the same modality). This shows the generality of spatial training for conceptual generalisation. This means that an explicit representation of space might be the substrate for strong forms of transfer observed in humans, such as when we understand the shared meaning between “red, amber, green” at a traffic light and “ready, set, go” before a race. We did not test transfer across modalities nor transfer in a more natural setting; we leave this for future studies.